# Techniques of Integration (Indefinite Integrals)

A regulatory scientist is staring at three rate equations on her screen — first-order pollutant decay with a constant source term, photosynthesis rate as a function of light that's itself a function of time, and a logistic-style population dynamic that just got fitted to a salmon counter dataset. Each one needs to be integrated to answer the actual question her agency cares about: *how much, in total, between these two dates?* The Fundamental Theorem of Calculus tells her there's an antiderivative out there for each rate. The hard part is *finding* it.

This chapter is about the toolkit for that hard part. The Fundamental Theorem reduces a definite integral to evaluating an antiderivative — but only if you can produce one. That isn't always obvious, because integration doesn't have a clean step-by-step algorithm the way differentiation does. What you build instead is a set of **pattern-recognition skills**: spotting when an integrand wants substitution, when it wants integration by parts, when it wants partial fractions, and when it really just wants to be rewritten before you do anything fancy at all.

## Purpose of the Chapter

In the previous chapter, you learned what a definite integral *means* (accumulation) and why it can be computed using antiderivatives (the Fundamental Theorem of Calculus).  

This chapter answers the next practical question:

> **How do we actually find antiderivatives when the integrand is not “obvious”?**

Rather than treating integration as a list of formulas, this chapter emphasizes **pattern recognition + strategy**: identify the structure of a function, choose an appropriate method, and interpret the result in context.

By the end of this chapter, you should be able to:

- compute **indefinite integrals** using core rules and techniques,
- recognize when to use **substitution**, **integration by parts**, **partial fractions**, or **numerical methods**,
- connect each technique back to derivative rules (chain rule, product rule),
- interpret results using **units, sign, and context**, not just algebra.

---

## Indefinite Integrals and the Constant of Integration

In the previous chapter, you learned how a **definite integral** measures *net accumulated change over a specific interval*. The bounds told you exactly where accumulation started and stopped, and the result was a **single number** with a clear physical meaning.

In this section, we remove those bounds and ask a different—but closely related—question:

> **What functions have a given rate of change?**

The answer leads us to **indefinite integrals**, antiderivatives, and the essential role of the constant of integration.

---

### From Definite to Indefinite Integrals

A **definite integral**
\[
\int_a^b f(x)\,dx
\]
asks:

> *How much total change occurred between \(x=a\) and \(x=b\)?*

An **indefinite integral**
\[
\int f(x)\,dx
\]
asks something more open-ended:

> *What function (or functions) have derivative equal to \(f(x)\)?*

When the bounds disappear, two important things change:

1. **The result is no longer a number.**  
   An indefinite integral produces a *function*.

2. **The accumulation is no longer anchored.**  
   Without a starting point, we cannot say how much has accumulated—only *how accumulation grows*.

This is why we write:
\[
\int f(x)\,dx = F(x) + C,
\]
where \(F'(x) = f(x)\).

The indefinite integral does not describe a single accumulated quantity.  
It describes an entire **family of possible accumulation functions**.

---

### Integrals as Families of Functions

Consider the rate function:
\[
f(x) = 2x.
\]

Any of the following functions has derivative \(2x\):
\[
x^2,\quad x^2 + 3,\quad x^2 - 10,\quad x^2 + C.
\]

```{r antiderivative-family-example, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}

library(ggplot2)
library(patchwork)

# Domain
x <- seq(-3, 3, length.out = 400)

# Rate function
f <- function(x) 2 * x

# Antiderivatives with different constants
F_base  <- x^2
F_up    <- x^2 + 3
F_down  <- x^2 - 10

# Data frames
df_rate <- data.frame(
  x = x,
  y = f(x)
)

df_anti <- data.frame(
  x = x,
  Base = F_base,
  ShiftUp = F_up,
  ShiftDown = F_down
)

# Left panel: rate function
p_rate <- ggplot(df_rate, aes(x = x, y = y)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Rate Function  f(x) = 2x",
    x = "x",
    y = "Rate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# Right panel: family of antiderivatives
p_anti <- ggplot(df_anti, aes(x = x)) +
  geom_line(aes(y = Base), linewidth = 1.2, color = "#0072B2") +
  geom_line(aes(y = ShiftUp), linewidth = 1.2, color = "#D55E00") +
  geom_line(aes(y = ShiftDown), linewidth = 1.2, color = "#009E73") +
  labs(
    title = "Antiderivatives  F(x) = x² + C",
    x = "x",
    y = "Accumulated Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# Combine panels
p_rate | p_anti

```

> Remember from Calc I that a derivative function f(x) describes the slopes of a function. The slopes of the functions on the right are all the same for the same x value. So a constant doesn't impact slope. 

All of these functions increase at the same rate for the same values of \(x\).  
The only difference is **where they start**.

This is the key conceptual shift:

> **An indefinite integral describes how accumulation changes, not how much has accumulated so far.**

The missing information is the starting value—and that missing information is encoded in the constant \(C\).

---

### Antiderivatives and the “+ C”

An **antiderivative** of \(f(x)\) is any function whose derivative is \(f(x)\).

Because derivatives eliminate constants:

Differentiate term by term:
\[
\frac{d}{dx}\big(F(x) + C\big)
= \frac{d}{dx}F(x) + \frac{d}{dx}C.
\]

Since a constant does not change with \(x\),
\[
\frac{d}{dx}C = 0.
\]

Therefore,
\[
\frac{d}{dx}\big(F(x) + C\big)
= F'(x) + 0
= F'(x).
\]

Every antiderivative comes with an unavoidable ambiguity.

That ambiguity is not a flaw—it reflects reality.

---

#### Why Antiderivatives Are Not Unique

If \(f(x)\) represents a **rate** (such as carbon flux, water inflow, or population growth), then the *accumulated amount* depends on **initial conditions**:

- How much carbon was already present?
- How much water was in the reservoir at the start?
- What was the initial population size?

Different starting values produce different accumulation curves—but all follow the **same rate of change**.

Mathematically, this is why we must include the constant of integration, \(+C\).

---

::: promptbox

**Environmental Example: Two Lakes with the Same Inflow Rate**

Suppose two lakes experience the **same inflow rate** of water, modeled by
\[
f(t) = 2t \quad \text{(m³/day)},
\]
where \(t\) is time in days.

This rate describes how quickly water is entering each lake at any moment.

Integrating the rate gives an accumulation function:
\[
V(t) = t^2 + C,
\]
where \(V(t)\) is the volume of water in the lake and \(C\) represents the **initial volume**.

Now suppose:

- **Lake A** starts with 100 m³ of water  
- **Lake B** starts with 250 m³ of water  

Both lakes gain water according to the *same rate*, so their accumulation functions have the same shape.  
The only difference is their starting point.

- Lake A: \(V_A(t) = t^2 + 100\)
- Lake B: \(V_B(t) = t^2 + 250\)

Graphically, these curves are **vertical shifts** of one another.

```{r two-lakes-same-rate-different-start, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}

library(ggplot2)

# Time domain (days)
t <- seq(0, 10, length.out = 400)

# Lake volume functions
VA <- t^2 + 100   # Lake A
VB <- t^2 + 250   # Lake B

# Data frame
df <- data.frame(
  t = t,
  LakeA = VA,
  LakeB = VB
)

# Plot
ggplot(df, aes(x = t)) +
  geom_line(aes(y = LakeA, color = "Lake A"), linewidth = 1.2) +
  geom_line(aes(y = LakeB, color = "Lake B"), linewidth = 1.2) +
  scale_y_continuous(limits = c(0, 400)) +
  labs(
    title = "Two Lakes with the Same Inflow Rate",
    subtitle = "Same rate of change, different initial volumes",
    x = "Time (days)",
    y = "Water Volume (m3)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )

```
**Key Insight**

> The rate of change determines the *shape* of accumulation.  
> Initial conditions determine the *vertical position*.

This is why antiderivatives are not unique:  
the mathematics reflects the physical reality that systems can evolve at the same rate while starting from different states.

:::


#### Vertical Shifts and Initial Conditions

The constant \(C\) represents a **vertical shift** of the accumulation function.

Conceptually:

- The *shape* of the curve is determined by the rate \(f(x)\).
- The *vertical position* is determined by initial conditions.

This mirrors environmental modeling:

- Same rainfall rate → same growth pattern of water volume  
- Different initial storage → different total volume at all times

Ignoring \(C\) is equivalent to assuming the system always starts at zero—which is rarely justified unless explicitly stated.

---

#### Why Constants Cancel in Definite Integrals

When computing a definite integral using an antiderivative:
\[
\int_a^b f(x)\,dx = [F(x) + C]_a^b,
\]
the constant cancels:
\[
(F(b) + C) - (F(a) + C) = F(b) - F(a).
\]


This is why definite integrals do **not** require a constant of integration.

A definite integral measures **change between two points**, not absolute level.  
Only differences matter.

Indefinite integrals, by contrast, describe **absolute accumulation**, so the constant matters.



---

### Quick Interpretation Habit

To use indefinite integrals correctly—especially in applied contexts—it helps to adopt two quick interpretation checks.

---

#### Units in Indefinite Integrals

Even without bounds, units still carry meaning.

If:

- \(f(x)\) has units of *rate* (e.g., kg/day),
- and \(x\) has units of *time* (days),

then:
\[
\int f(x)\,dx
\]
has units of **accumulated quantity** (kg).

However, without bounds:

- you do **not** yet know *how much* has accumulated,
- only how accumulation depends on \(x\).

This distinction prevents a common mistake: treating an indefinite integral as a physical total without specifying a reference point.

---

#### Determining \(C\) from a Known Starting Value

In real problems, the constant \(C\) is determined using **initial conditions**.

Example:
- Suppose \(f(t)\) is a rate of change.
- You find an antiderivative \(F(t) + C\).
- You are told that the accumulated quantity equals \(A_0\) at time \(t=t_0\).

Then you solve:
\[
F(t_0) + C = A_0
\]
to determine \(C\).

This step turns a *family* of solutions into a **specific model**.

---

### Key Takeaway

> **Indefinite integrals describe how accumulation grows; the constant of integration encodes where accumulation starts.**

Definite integrals eliminate this ambiguity by focusing on change between two points.  
Indefinite integrals retain it—because real systems depend on initial conditions.

Understanding this distinction is essential before learning any integration technique, because every technique ultimately produces an antiderivative—and every antiderivative represents a family of possible accumulated states.


---

## Reverse Differentiation as a Skill

In environmental science you usually know a *rate* — how fast a contaminant is decaying, how fast a forest stand is fixing carbon, how fast a population is changing — and you need to recover the *quantity* that produced that rate. That recovery is exactly antidifferentiation. The good news is that most of the rates you'll meet in introductory environmental modeling fall into a short list of recognizable shapes: power-law growth or decay, exponential approach to equilibrium, logarithmic accumulation, and the occasional periodic signal. Building fluency with a small **antiderivative dictionary** covers most of what you'll see.


At this stage, integration should not feel like a new topic—it should feel like **recognition**.

Every integration rule you learn already exists somewhere in your derivative toolkit. The challenge is not memorizing formulas, but training yourself to *see* an integrand as the derivative of something familiar.

Think of this section as building an **antiderivative dictionary** and learning how to look up entries quickly.

---

### The “Antiderivative Dictionary”

An antiderivative dictionary is a mental catalog that pairs common derivatives with their original functions. When you integrate, you are asking:

> *Have I seen this derivative before?*

Below are the core entries you should know cold.

---

#### Power Rule (in Reverse)

From differentiation:
\[
\frac{d}{dx}(x^n) = nx^{n-1}
\]

Reversing this:
\[
\int x^n \, dx = \frac{x^{n+1}}{n+1} + C \quad (n \neq -1)
\]

**Examples**
\[
\int x^4 dx = \frac{x^5}{5} + C
\]

\[
\int x^{-3} dx = \frac{x^{-2}}{-2} + C = -\frac{1}{2x^2} + C
\]

---

#### The Special Case: \(x^{-1}\)

The power rule fails when \(n = -1\).

Instead:
\[
\int \frac{1}{x}\,dx = \ln|x| + C
\]

This is not arbitrary—it comes directly from:
\[
\frac{d}{dx}(\ln|x|) = \frac{1}{x}
\]

**Key recognition trigger:**  
Any time you see a lone \(1/x\), think *logarithm*.

---

#### Exponentials and Logarithms

Exponentials are special because they reproduce themselves under differentiation.

\[
\frac{d}{dx}(e^x) = e^x
\]

So:
\[
\int e^x dx = e^x + C
\]

Similarly:
\[
\int a^x dx = \frac{a^x}{\ln a} + C \quad (a > 0, a \neq 1)
\]

**Logarithms**
\[
\frac{d}{dx}(\ln|x|) = \frac{1}{x}
\]

So:
\[
\int \frac{1}{x} dx = \ln|x| + C
\]

---

#### Basic Trigonometric Antiderivatives

These come directly from familiar derivative rules.

| Integrand | Antiderivative |
|:---------:|:---------------:|
| \(\cos x\) | \(\sin x + C\) |
| \(\sin x\) | \(-\cos x + C\) |
| \(\sec^2 x\) | \(\tan x + C\) |
| \(\csc^2 x\) | \(-\cot x + C\) |
| \(\sec x \tan x\) | \(\sec x + C\) |
| \(\csc x \cot x\) | \(-\csc x + C\) |

**Sign matters.**  
For sine and cosine in particular, the negative sign is part of the recognition.

---

### Pattern Recognition Over Memorization

A common misconception is that integration is about remembering formulas.

In reality, successful integration looks like this:

1. **Scan the integrand**
2. **Ask:** “What derivative does this resemble?”
3. **Match structure**, not symbols
4. **Apply the corresponding reverse rule**

---

#### Reading an Integrand as a Derivative

Consider:
\[
\int 6x^5 dx
\]

Instead of thinking “apply the power rule,” think:

> *What function has derivative \(6x^5\)?*

You know:
\[
\frac{d}{dx}(x^6) = 6x^5
\]

So immediately:
\[
\int 6x^5 dx = x^6 + C
\]

This mindset becomes essential later when functions are more complicated.

---

#### Common Structures That Should Trigger a Rule

Here are some quick recognition cues:

- **Powers of \(x\)** → power rule  
- **\(1/x\)** → logarithm  
- **\(e^x\)** → exponential  
- **Trig functions alone** → basic trig antiderivatives  
- **Constant multiples** → pull out the constant  

If nothing jumps out immediately, that’s a signal—not a failure. It often means a **strategy** (like substitution) is coming later.

---



### Key Takeaway

> **Integration is reverse differentiation plus pattern recognition.**

Build a small, reliable antiderivative dictionary.  
Practice recognizing structures before computing.  
Speed comes from understanding, not memorization.

Once this skill is in place, integration techniques become tools—not obstacles.

---

### Concept Check: Why Integration Is Harder Than Differentiation {-}

Differentiation has a step-by-step algorithm — power rule, chain rule, product rule, and you can grind through anything. Integration doesn't work that way. Why?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Differentiation is a **constructive** process: every elementary function has a derivative that can be built mechanically from a small set of rules. Integration is the *inverse* — you're trying to find a function whose derivative is given — and there's no general recipe that always works. Many elementary-looking integrands (like \(e^{-x^2}\) or \(\sin(x)/x\)) don't even have elementary antiderivatives at all. So instead of an algorithm, you learn to **recognize patterns** and pick a technique that matches.

</details>


### Concept Check: Reading an Integrand as a Derivative {-}

When you spot \(\int 2x\cos(x^2)\,dx\) and immediately think "that's a chain-rule derivative in disguise", what specifically are you noticing?

<details>
<summary><strong>Click to reveal answer</strong></summary>

You're noticing that one factor (\(2x\)) is **exactly the derivative** of the inner part of the other factor (the \(x^2\) inside the cosine). Differentiating \(\sin(x^2)\) by the chain rule would give \(\cos(x^2)\cdot 2x\) — exactly your integrand. So an antiderivative is \(\sin(x^2) + C\). The pattern that *triggers* this recognition is "inner function and its derivative are both present".

</details>


---

## Substitution: When a Function Is Hiding Inside Another

Substitution is the technique that handles **composite rates** — rates that depend on a quantity that is itself changing. This comes up constantly in environmental work: a chemical reaction rate that depends on concentration which depends on time, an oxygen-uptake rate that depends on temperature which depends on depth, a discharge rate written in one variable that needs to be re-expressed in another. Whenever an integrand contains a function tucked inside another function — and the derivative of that inner function happens to be sitting right there — substitution is your first move.


Many integrals fail not because the function is complicated, but because it is **nested**.

Substitution is the technique that lets us *temporarily ignore the outer complexity* by focusing on the inner structure. Conceptually, it is nothing more than the **chain rule in reverse**.

---

### The Big Idea: The Chain Rule in Reverse

Recall the chain rule from differentiation:
\[
\frac{d}{dx}\big(F(g(x))\big) = F'(g(x))\,g'(x).
\]

Substitution works when an integrand matches this derivative structure:
\[
F'(g(x))\,g'(x).
\]

For notational simplicity, we often write this pattern as
\[
f(g(x))\,g'(x),
\]
where \(f\) represents the derivative of some outer function \(F\).

The key recognition step is:

> **Do I see a function inside another function, along with (almost) its derivative?**

At first, this pattern can be difficult to spot. That is normal.  
Recognition improves with repetition, not memorization.

Equally important, we will practice **course correction**: learning how to recognize when substitution is *not* working and what that tells you about which strategy to try next.

---

#### Recognizing \(f(g(x))g'(x)\)

Consider:
\[
\int (3x^2 + 1)^5 \cdot 6x \, dx.
\]

Read it structurally:

- Inner function: \(g(x) = 3x^2 + 1\)
- Outer function: \(F(u) = u^5\)
- Derivative of the inner function: \(g'(x) = 6x\)

This matches the chain rule pattern exactly.

Instead of expanding powers or managing products, we *rename the inner function* and simplify the structure.

---

#### Why Substitution Is a Change of Variable (Not a Trick)

Substitution is often taught procedurally, which makes it feel like a trick.

Conceptually, it is much simpler:

> **We change variables so that the integral matches something we already know how to do.**

We are not changing the mathematics—only the *perspective*.

This mirrors modeling practice:

- If a process depends on temperature, we might re-express time-dependent data in terms of temperature.
- If a rate depends on concentration, we may analyze accumulation with concentration as the variable.

Substitution is exactly this idea, applied mathematically.

---

### The Mechanics of \(u\)-Substitution

Once the structure is recognized, the mechanics follow a consistent pattern.

---

#### Step 1: Choose \(u\)

Choose \(u\) to be the **inner function**.

Example:
\[
\int (3x^2 + 1)^5 \cdot 6x \, dx
\]

Let:
\[
u = 3x^2 + 1.
\]

---

#### Step 2: Convert \(dx\) to \(du\)

Differentiate \(u\):
\[
\frac{du}{dx} = 6x
\quad \Rightarrow \quad
du = 6x\,dx.
\]

This step is not symbolic bookkeeping—it is where the **meaning of the integral is preserved**.

---

##### What \(dx\) Actually Means

Recall where integrals come from.

A definite integral begins as a Riemann sum:
\[
\sum f(x_i)\,\Delta x.
\]

Here:
- \(f(x_i)\) is the height of each rectangle,
- \(\Delta x\) is the **width** of each rectangle.

When we take the limit as the widths shrink,
\[
\Delta x \to 0,
\]
the sum becomes an integral:
\[
\int f(x)\,dx.
\]

The symbol \(dx\) represents an **infinitesimally small width measured in the \(x\)-direction**.

---

##### Why \(dx\) Must Change When the Variable Changes

When we use substitution, we are no longer measuring accumulation in terms of \(x\).  
We are changing variables and measuring accumulation in terms of a new quantity \(u\).

That means:
- the “height” of each rectangle is rewritten in terms of \(u\),
- **the width must also be rewritten**.

Leaving \(dx\) behind would mix measurements from two different coordinate systems.

---

##### What \(du = 6x\,dx\) Is Really Saying

When we compute:
\[
du = 6x\,dx,
\]
we are expressing how an **infinitesimal change in \(x\)** corresponds to an **infinitesimal change in \(u\)**.

- \(dx\) measures tiny steps in \(x\),
- \(du\) measures the corresponding tiny steps in \(u\),
- the factor \(6x\) tells us how those widths scale.

The product \(6x\,dx\) is the **new width** after changing variables.

---

##### The Key Idea

> Substitution changes both the function *and* the width over which accumulation happens.

Failing to convert the differential breaks the meaning of the integral.

---

##### A Useful Mental Model

Think of substitution as re-measuring the same total accumulation using a **different ruler**.

- \(dx\) measures length in \(x\)-units,
- \(du\) measures length in \(u\)-units,
- \(\frac{du}{dx}\) tells you how to convert between them.

---

#### Step 3: Rewrite the Integral in Terms of \(u\)

Substitute:

- \((3x^2 + 1) \rightarrow u\)
- \(6x\,dx \rightarrow du\)

The integral becomes:
\[
\int u^5 \, du.
\]

Now the structure is simple.

---

#### Step 4: Integrate and Substitute Back

Integrate:
\[
\int u^5 du = \frac{u^6}{6} + C.
\]

Substitute back:
\[
\frac{(3x^2 + 1)^6}{6} + C.
\]

---

### Why Substitution Works

Substitution temporarily turns a **composite rate** into a **simple rate**.

You are effectively saying:

> *Let me measure accumulation with respect to the inner process first.*

Once that accumulation is understood, you translate it back to the original variable.

---

### Substitution in Definite Integrals

Substitution works the same way for definite integrals, with one improvement:  
you can **change the bounds instead of substituting back**.

---

#### Changing the Bounds

Example:
\[
\int_0^2 (3x^2 + 1)^5 \cdot 6x \, dx
\]

Let:
\[
u = 3x^2 + 1.
\]

Convert the bounds:

- When \(x = 0\): \(u = 1\)
- When \(x = 2\): \(u = 13\)

Now the integral becomes:
\[
\int_1^{13} u^5 \, du.
\]

---

#### Why This Reduces Errors

- No need to rewrite \(x\) in terms of \(u\)
- Cleaner algebra
- Fewer opportunities to lose constants or signs

---

::: promptbox

**Worked Example: Why the Differential Must Change**

Let’s look at a substitution example where the role of \(dx\) and \(du\) is especially clear.

---

Evaluate:
\[
\int 2x \, e^{x^2} \, dx
\]

---

**Step 1: Recognize the Structure**

- Inner function: \(x^2\)
- Outer function: \(F(u) = e^u\)
- Derivative of the inner function: \(2x\)

This matches the chain rule pattern.

---

**Step 2: Choose \(u\)**

Let:
\[
u = x^2.
\]

Then:
\[
du = 2x\,dx.
\]

👉 **The entire factor \(2x\,dx\) is exactly \(du\).**

---

**Step 3: Rewrite the Integral in Terms of \(u\)**

\[
\int e^{u}\,du.
\]

---

**Step 4: Integrate and Substitute Back**

\[
e^{x^2} + C.
\]

---

**Key Insight from This Example**

> The differential tells you **what variable the accumulation is happening in**.

Both the function and the width must be expressed in the same variable.

:::

---

### Common Pitfalls

Substitution errors almost always fall into one of the following categories.

---

#### Missing the Derivative Factor

\[
\int (x^2 + 1)^4 dx
\]

There is no \(2x\) present.

This means substitution will not simplify the integral without additional manipulation.

---

#### Incorrect \(du\)

\[
u = x^2 + 1 \quad \Rightarrow \quad du = 2x
\]

This is incomplete.

Correct:
\[
du = 2x\,dx.
\]

---

#### Substituting Back Too Early

Once the integral is rewritten in terms of \(u\), **stay in \(u\)** until after integration.

---

### Strategy Check

Before using substitution, ask:

- Do I see a nested structure?
- Is the derivative of the inner function present (up to a constant)?
- Will this rewrite the integral into something simpler?

If the answer is no, that is information—not failure.

---

### Key Takeaway

> **Substitution is the chain rule in reverse, framed as a change of variable.**

It simplifies nested structure by re-measuring accumulation in the variable that controls change.

If substitution fails, it usually fails clearly—telling you it is time for a different strategy.

---

### When the Derivative Is *Almost* There – *Close Enough Can Be Enough*

Very often, the derivative of the inner function is present **up to a constant factor**.

This is not a problem—it is a feature.

---

#### Example

Evaluate:
\[
\int (3x^2 + 1)^4 \cdot x \, dx
\]

---

#### Step 1: Recognize the Structure

- Inner function: \(3x^2 + 1\)
- Derivative: \(6x\)

The integrand contains \(x\), not \(6x\).

---

#### Step 2: Adjust with a Constant

\[
= \int (3x^2 + 1)^4 \cdot \frac{1}{6}(6x)\, dx
= \frac{1}{6} \int (3x^2 + 1)^4 \cdot 6x\, dx.
\]

---

#### Step 3: Substitute and Integrate

\[
\frac{1}{6} \int u^4 du
= \frac{(3x^2 + 1)^5}{30} + C.
\]

---

### Strategy Rule: *Course Correct*

> **Substitution tolerates constant mismatches—but not structural ones.**

If the mismatch depends on \(x\), substitution is no longer the right tool.

---

### Concept Check: Spotting When Substitution Is the Right Move {-}

What two features must an integrand have for substitution to work cleanly?

<details>
<summary><strong>Click to reveal answer</strong></summary>

1. **An inner function** — some quantity \(g(x)\) that's wrapped inside another function (like the \(x^2\) inside \(\cos(x^2)\), or the \(3x^2 + 1\) inside \((\,\cdot\,)^5\)).
2. **Its derivative is also present in the integrand** — \(g'(x)\) is sitting there as a factor, possibly up to a constant.

If both features are present, picking \(u = g(x)\) and converting \(dx\) to \(du\) collapses the composite structure into a simple integral. If the derivative is *missing entirely* (not just off by a constant), substitution won't help and you need a different technique.

</details>


### Concept Check: Why dx Must Become du, Not Just Disappear {-}

When you do a substitution, why isn't it enough to just rename \(x\) to \(u\) inside the function — why does \(dx\) also have to be converted to \(du\)?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Because \(dx\) and \(du\) measure widths in *different units*. The original integral \(\int f(x)\,dx\) is a limit of Riemann sums \(\sum f(x_i)\,\Delta x\) where \(\Delta x\) is the width of each slice **measured in \(x\)**. When you switch to \(u\), the slice widths are now **measured in \(u\)** — and \(\Delta u \neq \Delta x\) unless you account for the stretching factor \(du/dx = g'(x)\). The conversion \(du = g'(x)\,dx\) is the bookkeeping that keeps the accumulation honest.

</details>


---

## Integration by Parts: When Accumulation Involves a Product

Integration by Parts is the right tool when accumulation involves a **product of two functions**, and at least one of them gets simpler when you differentiate it. In environmental modeling this shows up whenever you're accumulating a quantity that is itself the product of two changing factors — total pollutant load (concentration × flow rate), total respiratory cost (per-capita rate × population size), heat lost over time (temperature difference × cooling rate). Each of these is naturally an integral of the form \(\int u\,dv\).


Substitution helps when a function is **nested**.  
Integration by parts is the complementary idea: it helps when a function is a **product**.

This technique is not new mathematics. It is simply the **product rule, rearranged** to solve a different question.

---

### The Big Idea: The Product Rule in Reverse

Recall the product rule from differentiation:
\[
\frac{d}{dx}\big(u(x)v(x)\big)
=
u'(x)v(x) + u(x)v'(x).
\]

This rule tells us how the rate of change of a **product** is built from two interacting pieces:  
one changing while the other is held fixed, and vice versa.

Integration by parts asks the inverse question:

> *If a rate involves a product, can we undo it?*

---

To do this, we rearrange the product rule to isolate one of the product terms:
\[
u(x)v'(x)
=
\frac{d}{dx}\big(u(x)v(x)\big) - u'(x)v(x).
\]

This step is purely algebraic—but conceptually important.  
It expresses a **product-type rate** as the difference between:

- the derivative of the full product, and
- a new product that may be simpler.

---

Now integrate both sides:
\[
\int u(x)v'(x)\,dx
=
u(x)v(x) - \int u'(x)v(x)\,dx.
\]

This is **integration by parts**.

---

### What This Formula Is Really Doing

Integration by parts does not magically evaluate an integral.  
It **rearranges where the difficulty lives**.

- One factor is differentiated (often simplifying it),
- the other is integrated (often staying manageable),
- the remaining integral is easier than the original.

In other words:

> **We trade one product for another, hoping the trade is in our favor.**

If the new integral is simpler, the method succeeds.  
If it is not, that is useful information—it means a different strategy is needed.

---

### Why This Mirrors the Product Rule

Just as substitution reverses the chain rule,  
integration by parts reverses the product rule.

- Substitution handles *nesting*.
- Integration by parts handles *interaction*.

Recognizing which structure you are facing is the key skill—not memorizing formulas.


### The Formula (and What It Means)

We usually write the result as:
\[
\int u\,dv = uv - \int v\,du.
\]

This is not something to memorize blindly.  
It is a bookkeeping statement that encodes the product rule.

Conceptually:

> **We trade one product for another that is easier to integrate.**

---

### When Integration by Parts Is the Right Tool

Integration by parts is useful when:

- the integrand is a **product** of two functions, and
- differentiating one factor simplifies it.

Classic triggers include:

- polynomials multiplied by exponentials or trig functions,
- logarithms multiplied by anything,
- inverse trig functions multiplied by powers of \(x\).

A useful diagnostic question is:

> **If I differentiate one factor, does it become simpler?**

If yes, integration by parts is a good candidate.

::: promptbox

**Quick Diagnostic Example**

Consider the integral:
\[
\int x \sin x \, dx.
\]

Before doing any computation, ask:

> *If I differentiate one factor, does it become simpler?*

- Differentiating \(x\) gives \(1\), which is simpler.
- Differentiating \(\sin x\) gives \(\cos x\), which is not simpler in any meaningful way.

This suggests choosing \(u = x\) is a good strategic move.

---

**What a “Good” Choice Looks Like**

- Start with: \(x \sin x\)
- Differentiate \(x \rightarrow 1\)

The product has lost structure.  
That is what “easier” means here.

---

**What a “Bad” Choice Looks Like**

If instead you differentiate the other factor:

- Start with: \(x \sin x\)
- Differentiate \(\sin x \rightarrow \cos x\)

The product is still a product of two nontrivial functions.  
Nothing has simplified.

This is a warning sign.

---

**The Takeaway**

“Easier” does not mean *finished*.  
It means **less structure than before**.

If differentiating a factor removes complexity, integration by parts is likely to help.

:::


---

### Choosing \(u\) and \(dv\)

The method requires a choice:

- one part becomes \(u\),
- the other becomes \(dv\).

There is no single correct rule, but a common guideline is:

> Choose \(u\) to be the factor that becomes simpler when differentiated.

This often—but not always—means:

- logarithms,
- inverse trig functions,
- algebraic powers of \(x\).

---

### A First Worked Example

Evaluate:
\[
\int x e^x \, dx.
\]

---

#### Step 1: Choose \(u\) and \(dv\)

Let:
\[
u = x.
\]

Take the derivative of both sides with respect to \(x\):
\[
\frac{du}{dx} = \frac{d}{dx}(x) = 1.
\]

Multiply both sides by \(dx\):
\[
du = dx.
\]

This tells us that an infinitesimal change in \(u\) is the same as an infinitesimal change in \(x\).

---

Now choose the remaining part of the integrand to be \(dv\):

Let:
\[
dv = e^x\,dx.
\]

To find \(v\), integrate both sides:
\[
\int dv = \int e^x\,dx.
\]

This gives:
\[
v = e^x.
\]

Here, we are identifying which part of the product will be **accumulated** (integrated) when we reverse the product rule.

---

Together, these steps do the following:

- \(u\) is the factor we plan to **differentiate**, producing \(du\),
- \(dv\) is the factor we plan to **integrate**, producing \(v\).

This choice determines how the product rule is unwound in the next step.


---

#### Step 2: Apply the Formula

\[
\int u\,dv = uv - \int v\,du
\]

\[
\int x e^x dx = x e^x - \int e^x dx.
\]

---

#### Step 3: Finish the Integral

\[
x e^x - e^x + C.
\]

Or factor:
\[
e^x(x - 1) + C.
\]

---

### Why This Worked

Originally, the integral involved a product that could not be simplified directly.

By differentiating \(x\), we turned it into a constant.  
By integrating \(e^x\), we left it unchanged.

The result was an integral that was **simpler than the original**.

That is the goal of integration by parts.

---

### A Second Example: Logarithms

Evaluate:
\[
\int \ln x \, dx.
\]

At first glance, this looks like it should require substitution—but there is no obvious inner function.

Instead, think of it as a product:
\[
\ln x = (\ln x)\cdot 1.
\]

---

#### Step 1: Choose \(u\) and \(dv\)

Let:
\[
u = \ln x
\quad \Rightarrow \quad
du = \frac{1}{x} dx.
\]

Let:
\[
dv = dx
\quad \Rightarrow \quad
v = x.
\]

---

#### Step 2: Apply the Formula

\[
\int \ln x \, dx
= x\ln x - \int x \cdot \frac{1}{x} dx.
\]

---

#### Step 3: Simplify

\[
x\ln x - \int 1\,dx
= x\ln x - x + C.
\]

---

### A Useful Interpretation

Integration by parts redistributes *where the complexity lives*.

- One factor is differentiated (losing complexity),
- the other is integrated (often staying similar).

If the trade makes the remaining integral simpler, the method succeeds.  
If it does not, that is a signal to try a different strategy.

---

### Common Pitfalls

---

#### Choosing \(u\) That Gets More Complicated

Example:
\[
\int x e^x dx
\]

Choosing:
\[
u = e^x
\]
does not help—its derivative is still \(e^x\).

Good choices simplify when differentiated.

---

#### Forgetting the Minus Sign

The formula is:
\[
\int u\,dv = uv - \int v\,du.
\]

The minus sign matters and is a common source of errors.

---

#### Stopping Too Soon

Sometimes, integration by parts must be applied **more than once**.

Example:
\[
\int x^2 e^x dx
\]

Each application reduces the power of \(x\) until the integral becomes straightforward.

---

### Strategy Check

Before using integration by parts, ask:

- Is the integrand a product?
- Will differentiating one factor simplify it?
- Does substitution clearly *not* apply?

If yes, integration by parts is likely the right tool.

---

### Key Takeaway

> **Integration by parts is the product rule in reverse.**

It works by trading one product for another that is easier to integrate.

Mastery comes not from memorizing the formula, but from recognizing when a product structure can be simplified through differentiation.

Once this technique is comfortable, you are ready to handle integrals involving products, logarithms, and repeated structure—especially those that arise naturally in modeling and applications.

---

### Concept Check: Choosing u and dv in Integration by Parts {-}

When you apply integration by parts to \(\int u\,dv = uv - \int v\,du\), how should you decide which factor is \(u\) and which is \(dv\)?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Pick \(u\) to be the factor that **gets simpler when you differentiate it**. Polynomials shrink (\(x^2 \to 2x \to 2 \to 0\)). Logarithms turn into algebraic fractions (\(\ln x \to 1/x\)). Inverse-trig functions turn into algebraic fractions too. These are good \(u\) candidates.

The remaining factor becomes \(dv\), and you need to be able to integrate it cleanly to find \(v\). Exponentials and basic trig functions are good \(dv\) candidates because integrating them doesn't make them worse.

A common mnemonic: **LIATE** (Logarithmic, Inverse trig, Algebraic, Trig, Exponential) — earlier-letter functions are usually your \(u\).

</details>


### Concept Check: Why Integration by Parts Works {-}

What's the connection between integration by parts and the product rule for derivatives?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Integration by parts **is the product rule, integrated**. Start with \(\frac{d}{dx}(uv) = u'v + uv'\). Integrate both sides:
\[
uv = \int u'v\,dx + \int uv'\,dx.
\]
Rearrange:
\[
\int uv'\,dx = uv - \int u'v\,dx,
\]
which is exactly \(\int u\,dv = uv - \int v\,du\) in differential notation. It's not a separate technique — it's the product rule running backwards, the same way substitution is the chain rule running backwards.

</details>


---

## Partial Fractions: When a Rational Function Can Be Taken Apart

Some of the most important environmental rate equations — **logistic growth**, **Michaelis–Menten enzyme kinetics**, and various **two-compartment mixing models** — produce rational functions (polynomial over polynomial) once you separate variables and try to integrate. Partial fractions is the technique that splits those rational functions into pieces simple enough to integrate using the dictionary entries you already have. It is the workhorse behind a lot of mathematical ecology.


Substitution helps when a function is **nested**.  
Integration by parts helps when a function is a **product**.  

Partial fractions address a different structure altogether:  
**ratios of polynomials**.

When an integrand is a rational function,
\[
\frac{P(x)}{Q(x)},
\]
partial fractions allows us to rewrite that ratio as a **sum of simpler pieces** whose antiderivatives we already know.

---

### The Big Idea: Decomposing a Ratio into Simpler Parts

The core idea behind partial fractions is not calculus—it is algebra.

Instead of integrating a complicated fraction directly, we ask:

> *Can this rational function be written as a sum of simpler rational functions?*

If the answer is yes, then integration becomes straightforward, because integration is linear:
\[
\int (f(x) + g(x))\,dx = \int f(x)\,dx + \int g(x)\,dx.
\]

---

### When Partial Fractions Is the Right Tool

Partial fractions is appropriate when:

- the integrand is a **rational function** (a polynomial divided by a polynomial),
- the degree of the numerator is **less than** the degree of the denominator (or can be made so),
- the denominator can be **factored** into simpler pieces.

If these conditions are met, partial fractions is often the most direct strategy.

---

### Step 0: Make the Rational Function Proper

Before decomposing, always check degrees.

If:
\[
\deg(P) \ge \deg(Q),
\]
perform **polynomial division** first.

Example:
\[
\frac{x^2 + 1}{x - 1}
= x + 1 + \frac{2}{x - 1}.
\]

Only the proper fraction needs partial fractions.

---

### The Big Structural Insight

The denominator controls the form of the decomposition.

Each factor in the denominator corresponds to a term in the decomposition.

We match structure first—just as with substitution and integration by parts.

---

#### Why We Break the Fraction Apart

At first glance, partial fractions can feel like we are making the problem *more* complicated by introducing multiple terms.

In reality, we are doing the opposite.

A rational function such as
\[
\frac{P(x)}{Q(x)}
\]
is often difficult to integrate directly because it does **not** match any antiderivative rule we know.

But integration is linear:
\[
\int (f(x) + g(x))\,dx = \int f(x)\,dx + \int g(x)\,dx.
\]

So our goal is to rewrite a single complicated ratio as a **sum of simple ratios**, each of which matches a familiar antiderivative.

---

#### What We Are Trying to Create

When the denominator factors into linear terms, we aim to produce fractions of the form:
\[
\frac{1}{x - a}.
\]

This is not arbitrary.

We do this because we already know the antiderivative:
\[
\int \frac{1}{x - a}\,dx = \ln|x - a| + C.
\]

Once the integrand is written in this form, the calculus is finished.

---

#### How the Denominator Dictates the Form

Each factor in the denominator represents a potential *rate behavior* that contributes to the overall expression.

For example:
\[
(x - 1)(x + 2)
\]
leads to terms of the form:
\[
\frac{A}{x - 1} + \frac{B}{x + 2}.
\]

Each term isolates the influence of one factor in the denominator and turns it into a rate with a known accumulation rule.

---

#### The Unifying Idea

Partial fractions is not about solving for constants.

It is about **engineering the integrand** so that every piece matches a known antiderivative identity.

The constants \(A\) and \(B\) are simply whatever values make the algebra work.

---

#### Structural Parallel

This mirrors earlier techniques:

- Substitution rewrites a complicated expression so it matches the chain rule in reverse.
- Integration by parts rewrites a product so it matches the product rule in reverse.
- Partial fractions rewrites a ratio so it matches **logarithmic and power-rule antiderivatives**.

In every case, the strategy is the same:

> **Change the form, not the meaning, until the structure becomes familiar.**

That is the real power of partial fractions.

---

### Case 1: Distinct Linear Factors

Consider:
\[
\int \frac{5x + 1}{(x - 1)(x + 2)}\,dx.
\]

Because the denominator factors into distinct linear terms, we write:
\[
\frac{5x + 1}{(x - 1)(x + 2)}
=
\frac{A}{x - 1} + \frac{B}{x + 2}.
\]

The goal is to find constants \(A\) and \(B\).

---

#### Solving for the Coefficients

Start with the decomposition:
\[
\frac{5x + 1}{(x - 1)(x + 2)}
=
\frac{A}{x - 1} + \frac{B}{x + 2}.
\]

Multiply both sides by the common denominator \((x - 1)(x + 2)\):
\[
5x + 1 = A(x + 2) + B(x - 1).
\]

Expand the right-hand side:
\[
5x + 1 = Ax + 2A + Bx - B.
\]

Combine like terms:
\[
5x + 1 = (A + B)x + (2A - B).
\]

Match coefficients:

- Coefficient of \(x\):
\[
A + B = 5
\]

- Constant term:
\[
2A - B = 1
\]

Solve the system.

Add the equations:
\[
(A + B) + (2A - B) = 5 + 1
\]
\[
3A = 6
\]
\[
A = 2.
\]

Substitute into \(A + B = 5\):
\[
2 + B = 5
\]
\[
B = 3.
\]

So the coefficients are:
\[
A = 2, \quad B = 3.
\]

---

#### Integrate Term by Term

Once the decomposition is complete, the integral becomes a sum of simpler integrals:
\[
\int \left(\frac{2}{x - 1} + \frac{3}{x + 2}\right)\,dx.
\]

Use linearity to split the integral:
\[
\int \left(\frac{2}{x - 1} + \frac{3}{x + 2}\right)\,dx
=
\int \frac{2}{x - 1}\,dx + \int \frac{3}{x + 2}\,dx.
\]

Now pull constants outside each integral:
\[
\int \frac{2}{x - 1}\,dx + \int \frac{3}{x + 2}\,dx
=
2\int \frac{1}{x - 1}\,dx + 3\int \frac{1}{x + 2}\,dx.
\]

Use the logarithm antiderivative identity:
\[
\int \frac{1}{x-a}\,dx = \ln|x-a| + C.
\]

Apply it to each term:

- For \(a=1\):
\[
2\int \frac{1}{x - 1}\,dx = 2\ln|x - 1|.
\]

- For \(a=-2\):
\[
3\int \frac{1}{x + 2}\,dx = 3\ln|x + 2|.
\]

Combine the results:
\[
2\ln|x - 1| + 3\ln|x + 2| + C.
\]

The single constant \(C\) at the end accounts for the constants that would arise from each integral.

---

### Case 2: Repeated Linear Factors

If the denominator contains a repeated linear factor such as
\[
(x - a)^2,
\]
then **each power of that factor requires its own term** in the decomposition.

This ensures that the decomposition has enough flexibility to match the original numerator.

---

#### Example

Evaluate:
\[
\int \frac{2x + 3}{(x - 1)^2}\,dx.
\]

This example is deliberately chosen so that **both terms in the decomposition matter**.

---

#### Step 1: Write the Partial Fraction Decomposition

Because the factor \((x - 1)\) appears with power 2, we must include **one term for each power**:
\[
\frac{2x + 3}{(x - 1)^2}
=
\frac{A}{x - 1} + \frac{B}{(x - 1)^2}.
\]

The goal is to determine the constants \(A\) and \(B\).

---

#### Step 2: Clear the Denominator

Multiply both sides by \((x - 1)^2\):
\[
2x + 3 = A(x - 1) + B.
\]

This equation must hold for **all** values of \(x\).

---

#### Step 3: Choose a Convenient Value of \(x\)

Because the equation is true for all \(x\), we can choose values that simplify the algebra.

---

##### Why Choose \(x = 1\)?

Substituting \(x = 1\) eliminates the term involving \(A\):
\[
2(1) + 3 = A(1 - 1) + B.
\]

So:
\[
5 = 0 + B,
\]
which gives
\[
B = 5.
\]

This isolates one coefficient immediately.

---

#### Step 4: Solve for the Remaining Coefficient

Substitute \(B = 5\) back into the equation:
\[
2x + 3 = A(x - 1) + 5.
\]

Subtract 5 from both sides:
\[
2x - 2 = A(x - 1).
\]

Factor the left-hand side:
\[
2(x - 1) = A(x - 1).
\]

Since this must hold for all \(x \neq 1\), we conclude:
\[
A = 2.
\]

---

#### Step 5: Write the Final Decomposition

\[
\frac{2x + 3}{(x - 1)^2}
=
\frac{2}{x - 1} + \frac{5}{(x - 1)^2}.
\]

This is the key payoff of partial fractions:  
a single complicated ratio rewritten as a sum of simpler pieces.

---

#### Step 6: Integrate Term by Term

Split the integral:
\[
\int \left(\frac{2}{x - 1} + \frac{5}{(x - 1)^2}\right)dx.
\]

Integrate each term separately:

\[
\int \frac{2}{x - 1}dx = 2\ln|x - 1|
\]

\[
\int \frac{5}{(x - 1)^2}dx
=
5\int (x - 1)^{-2}dx
=
-\frac{5}{x - 1}.
\]

Combine the results:
\[
2\ln|x - 1| - \frac{5}{x - 1} + C.
\]

---

### Key Insight

For repeated linear factors:

- **each power of the factor must appear in the decomposition**,  
- repeated factors produce **both logarithmic and power-rule terms**,  
- choosing values that zero out factors simplifies coefficient solving,  
- once decomposed, integration uses familiar rules only.

This case shows *why* repeated factors require extra terms:  
without them, the algebra cannot reproduce the original numerator.

As always, the strategy is:

> **Rewrite first. Integrate second.**


---

### Why Partial Fractions Works

Partial fractions succeeds because it converts a **single complicated rate** into a **sum of simple rates**.

Each piece corresponds to a familiar antiderivative:
- logarithms,
- power rules,
- inverse trigonometric functions.

Rather than inventing a new technique, partial fractions allows us to reuse old ones.

---

### Common Pitfalls

---

#### Skipping Polynomial Division

If the fraction is not proper, decomposition will fail or become messy.

Always divide first.

---

#### Missing Terms in the Decomposition

Each factor and each power must appear.

For example:
\[
(x - 1)^3 \Rightarrow \frac{A}{x - 1} + \frac{B}{(x - 1)^2} + \frac{C}{(x - 1)^3}.
\]

---

#### Forgetting the Goal

Partial fractions is not about finding \(A\), \(B\), and \(C\).  
It is about rewriting the integrand into **integrable pieces**.

---

### Strategy Check

Before using partial fractions, ask:

- Is this a rational function?
- Is the fraction proper (or can it be made proper)?
- Can the denominator be factored?

If yes, partial fractions is likely the right tool.

---

### Key Takeaway

> **Partial fractions replaces a difficult rational expression with a sum of simple ones.**

It works because integration is linear and familiar antiderivatives are reusable.

Once mastered, partial fractions becomes a powerful algebraic tool for turning complex ratios into manageable integrals—especially in applied models involving rational rates.


### Concept Check: Recognizing When Partial Fractions Is the Right Tool {-}

You see an integral \(\int \frac{P(x)}{Q(x)}\,dx\) where both \(P\) and \(Q\) are polynomials. What signals tell you partial fractions is the right move (versus substitution or something else)?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Partial fractions is the right move when **all three** of the following are true:

1. The integrand is a *ratio* of polynomials.
2. The degree of the numerator is *less than* the degree of the denominator (if not, do polynomial long division first).
3. There is **no obvious substitution** — i.e., the derivative of the denominator isn't sitting in the numerator (because if it were, you'd use substitution instead, since \(\int Q'(x)/Q(x)\,dx = \ln|Q(x)| + C\)).

When those line up, factor the denominator, decompose into pieces, and integrate each piece.

</details>


### Concept Check: Why the Decomposition Depends on the Denominator's Factors {-}

When you decompose a rational function into partial fractions, why does the form of the decomposition depend on *what kind* of factors are in the denominator (distinct linear, repeated linear, irreducible quadratic)?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Because each type of factor needs a different *kind* of antiderivative-friendly piece to be the inverse of differentiating it.

- A **distinct linear factor** \((x - a)\) corresponds to a piece \(\frac{A}{x-a}\), which integrates to a logarithm.
- A **repeated linear factor** \((x - a)^2\) needs *both* \(\frac{A}{x-a}\) **and** \(\frac{B}{(x-a)^2}\), because the second piece integrates to \(-1/(x-a)\) — a different shape — and you need both shapes available.
- An **irreducible quadratic factor** \((x^2 + bx + c)\) needs a numerator of the form \(Ax + B\), because that's what integrates cleanly into a mix of \(\ln\) and \(\arctan\) pieces.

The form of each piece is *engineered* to land on something you can integrate from the dictionary.

</details>


---

## Strategy: How to Choose a Method

By this point, you have seen several integration techniques.  
The challenge now is no longer *how* to execute a method, but **which method to try**—and in what order.

This section focuses on strategy: recognizing structure, choosing wisely, and avoiding common traps that come from committing too early.

---

### The Decision Tree

Every integration problem begins the same way:

> **Pause before computing.**

Ask what kind of object you are looking at.

---

#### Step 1: Look for the Simplest Case First

Always start by asking:

> *Can this be integrated directly using basic rules?*

Try this **before** anything else.

Basic rules include:

- Power rule
- Exponentials
- Logarithms
- Basic trigonometric antiderivatives
- Linearity (splitting sums, pulling out constants)

If the integrand is already a sum of familiar forms, you are done.

---

#### Step 2: Check for a Nested Structure → Substitution

If basic rules do not apply, ask:

> *Do I see a function inside another function, along with (almost) its derivative?*

Indicators for substitution:

- Powers of expressions like \((3x^2+1)^5\)
- Exponentials like \(e^{x^2}\)
- Rational expressions where the numerator resembles the derivative of the denominator

If the answer is yes, try substitution next.

---

#### Step 3: Check for a Product → Integration by Parts

If substitution does not simplify the integrand, ask:

> *Is the integrand a product of two functions?*

Then ask the diagnostic question:

> *If I differentiate one factor, does it become simpler?*

Common triggers:

- \(x e^x\)
- \(x \sin x\)
- \(\ln x\)
- inverse trig functions times algebraic factors

If differentiating one part reduces complexity, integration by parts is a good candidate.

---

#### Step 4: Check for a Rational Function → Partial Fractions

If the integrand is a ratio of polynomials:
\[
\frac{P(x)}{Q(x)},
\]
ask:

- Is the degree of the numerator less than the degree of the denominator?
- Can the denominator be factored?

If yes, partial fractions is often the right tool.

---

#### Step 5: When None of These Apply

If:

- the function does not simplify,
- no structure matches known rules,
- or the antiderivative cannot be written in elementary functions,

then **numerical methods** or approximation may be appropriate.

Not every integral has a closed-form solution—and recognizing that is part of mathematical maturity.

---

#### A Compact “What Do I Try First?” Checklist

1. Simplify algebraically (expand, cancel, split)
2. Apply basic rules if possible
3. Look for substitution
4. Look for integration by parts
5. Look for partial fractions
6. Consider numerical or approximate methods

Order matters. Skipping early steps creates unnecessary difficulty.

---

#### Strategy Is Reversible

Trying a method and backing out is not failure.

It is feedback.

If:

- substitution makes the expression worse,
- integration by parts increases complexity,
- partial fractions cannot be set up cleanly,

that information tells you something important about the structure of the problem.

---

### Workflow Habits That Prevent Mistakes

Good integration is not just about technique—it is about habits.

---

#### Rewrite and Simplify First

Before choosing a method:

- expand products if helpful,
- cancel factors,
- rewrite powers and fractions,
- split sums.

Many “hard” integrals become easy after algebraic cleanup.

---

#### Check by Differentiating

Whenever possible, differentiate your result.

Ask:

- *Do I recover the original integrand?*
- *Does the structure match?*

This is the fastest way to catch sign errors, missing constants, or incorrect substitutions.

---

#### Units and Sign Checks (Especially in Applications)

In applied contexts:

- Check that the units of the result make sense
- Check whether the sign matches the physical interpretation
  - accumulation should increase when rates are positive,
  - decreases should be reflected by negative values.

A mathematically correct answer that violates units or interpretation is still wrong.

---

### Final Strategy Takeaway

> **Integration is not about choosing the “right formula.”  
> It is about recognizing structure and making the problem simpler, step by step.**

When in doubt:

- simplify,
- match structure,
- choose the method that reduces complexity,
- and let failed attempts guide your next move.

That is how experts integrate—and how you will too.

---

### Concept Check: Always Try to Rewrite First {-}

Before reaching for substitution, IBP, or partial fractions, what should always be your first step on an integral that doesn't look obvious?

<details>
<summary><strong>Click to reveal answer</strong></summary>

**Rewrite and simplify.** Distribute products, split fractions, separate sums, cancel common factors, use trig identities, pull constants out front. A surprising number of "hard" integrals collapse into trivial dictionary applications once you stop and tidy up the integrand. Reaching for a heavy technique on a not-yet-simplified integrand is one of the most common ways to make a problem worse.

</details>


### Concept Check: Why Differentiation Is the Strongest Check {-}

After computing an indefinite integral, why is *differentiating your answer* the strongest way to check it — stronger than re-doing the integration?

<details>
<summary><strong>Click to reveal answer</strong></summary>

Because differentiation has a **deterministic algorithm**. If you differentiate your proposed antiderivative \(F(x)\) and don't get back exactly the integrand \(f(x)\), you know with certainty that something is wrong. Re-doing the integration can repeat the same mistake; differentiating the answer is an *independent* test that uses different machinery. A correct antiderivative differentiates back to the integrand every time, and the check takes about 30 seconds.

</details>


---

## Practice

### Practice: Fast Recognition Drills

These exercises are not about careful algebra. They are about **speed and confidence**.

The goal is to train your brain to identify patterns *before* computing.

---

#### Drill A: Name the Rule (No Computing) {-}

For each integrand, identify the rule you would use.

1. \(\int x^7 dx\)  
2. \(\int \frac{1}{x} dx\)  
3. \(\int e^x dx\)  
4. \(\int 4\cos x dx\)  
5. \(\int x^{-2} dx\)

*Do not integrate yet. Just name the rule.*

<details>
<summary><strong>Click to reveal solutions</strong></summary>

1. Power rule  
2. Logarithmic rule (\(\int 1/x\,dx\))  
3. Exponential rule  
4. Trigonometric rule (cosine) + constant multiple  
5. Power rule (negative exponent)

</details>

---

#### Drill B: One-Step Integrals {-}

Now compute quickly.

1. \(\int 3x^2 dx\)  
2. \(\int \sin x dx\)  
3. \(\int \frac{5}{x} dx\)  
4. \(\int e^x dx\)  
5. \(\int \cos x dx\)

<details>
<summary><strong>Click to reveal solutions</strong></summary>

1. \(x^3 + C\)  
2. \(-\cos x + C\)  
3. \(5\ln|x| + C\)  
4. \(e^x + C\)  
5. \(\sin x + C\)

</details>

---

#### Drill C: Recognition First, Then Solve {-}

Before integrating, say out loud:  
> *“This looks like the derivative of…”*

1. \(\int 8x^3 dx\)  
2. \(\int \cos x dx\)  
3. \(\int a^x dx\)

Then compute.

<details>
<summary><strong>Click to reveal solutions</strong></summary>

1. \(2x^4 + C\)  
   (since \(\frac{d}{dx}(2x^4) = 8x^3\))

2. \(\sin x + C\)  
   (since \(\frac{d}{dx}(\sin x) = \cos x\))

3. \(\frac{a^x}{\ln a} + C\)  
   (since \(\frac{d}{dx}(a^x) = a^x \ln a\))

</details>

---

#### Why These Drills Matter {-}

Later integration techniques—substitution, integration by parts, partial fractions—**all depend on recognition**.

If you cannot instantly recognize basic antiderivatives, advanced techniques will feel overwhelming.

Think of this section as learning vocabulary before writing sentences.

---

### Practice: Substitution (u-Sub) Pattern Recognition and Application

This practice section is designed to slow things down **before** computation.  
Your goal is to *see the structure* first, then execute the mechanics.

---

#### Part A: Identifying the Parts of the Integral (Pattern Recognition) {-}

For each integrand below:

1. Identify the **inner function** (if there is one).
2. Identify the **derivative of the inner function**.
3. Decide whether substitution is appropriate.
4. **Check whether the derivative is present exactly, off by a constant, or missing.**

Do **not** compute the integral yet.

---

#### A Helpful Instruction: “Off by a Constant” Is Still OK {-}

When checking for substitution, the derivative of the inner function does **not** need to appear perfectly.

If the integrand contains
\[
g'(x)\quad \text{or} \quad c\,g'(x),
\]
where \(c\) is a constant, substitution still works.

In that case:
- factor out the constant,
- proceed with substitution,
- and account for the constant in front of the integral.

What *does not* work is when the mismatch depends on \(x\).

---

#### Warning: Some Problems Are Designed *Not* to Use Substitution {-}

The list below intentionally mixes problems where substitution **does** work with problems where it **does not**.

Your job is to tell the difference *before* computing.

---

#### A1. {-}

\[
\int (3x^2 + 1)^5 \cdot 6x \, dx
\]

#### A2. {-}

\[
\int (x^2 + 1)^4 \, dx
\]

#### A3.{-}

\[
\int e^{x^2} \cdot 2x \, dx
\]

#### A4. {-}

\[
\int x e^{x} \, dx
\]

#### A5.{-}

\[
\int \frac{4x}{x^2 + 7} \, dx
\]

#### A6.{-}

\[
\int \cos(5x)\, dx
\]

#### A7.{-}

\[
\int (x + 1)^4 \, dx
\]

---

<details>
<summary><strong>Click to reveal analysis</strong></summary>

- **A1**
  - Inner function: \(3x^2 + 1\)  
  - Derivative: \(6x\)  
  - Match: exact  
  - Substitution: ✔️

- **A2**
  - Inner function: \(x^2 + 1\)  
  - Derivative: \(2x\)  
  - Match: missing  
  - Substitution: ✖️  
  - Reason: no factor of \(x\) to supply \(du\)

- **A3**
  - Inner function: \(x^2\)  
  - Derivative: \(2x\)  
  - Match: exact  
  - Substitution: ✔️

- **A4**
  - Structure: product, not nesting  
  - Substitution: ✖️  
  - Reason: derivative mismatch depends on \(x\); integration by parts is appropriate

- **A5**
  - Inner function: \(x^2 + 7\)  
  - Derivative: \(2x\)  
  - Match: off by a constant  
  - Substitution: ✔️

- **A6**
  - Inner function: \(5x\)  
  - Derivative: \(5\)  
  - Match: off by a constant  
  - Substitution: ✔️

- **A7**
  - Inner function: \(x + 1\)  
  - Derivative: \(1\)  
  - Match: exact  
  - Substitution: ✔️ (though basic power rule also works)

</details>

---

#### Part B: Solve Using Substitution {-}

Now compute each integral **where substitution is appropriate**.  
If substitution is *not* appropriate, do not force it.

---

#### B1.{-}

\[
\int (3x^2 + 1)^5 \cdot 6x \, dx
\]

#### B2.{-}

\[
\int e^{x^2} \cdot 2x \, dx
\]

#### B3.{-}

\[
\int \frac{4x}{x^2 + 7} \, dx
\]

#### B4.{-}

\[
\int \cos(5x)\, dx
\]

#### B5.{-}

\[
\int (x + 1)^4 \, dx
\]

---

<details>
<summary><strong>Click to reveal solutions</strong></summary>

- **B1**
  - Let \(u = 3x^2 + 1\), \(du = 6x\,dx\)
  \[
  \frac{(3x^2 + 1)^6}{6} + C
  \]

- **B2**
  - Let \(u = x^2\), \(du = 2x\,dx\)
  \[
  e^{x^2} + C
  \]

- **B3**
  - Let \(u = x^2 + 7\), \(du = 2x\,dx\)
  \[
  2\ln(x^2 + 7) + C
  \]

- **B4**
  - Let \(u = 5x\), \(du = 5\,dx\)
  \[
  \frac{1}{5}\sin(5x) + C
  \]

- **B5**
  - Let \(u = x + 1\), \(du = dx\)
  \[
  \frac{(x + 1)^5}{5} + C
  \]

</details>

---

#### Reflection Prompt {-}

After completing this section, ask yourself:

- Did I identify the inner function *before* choosing \(u\)?
- Did I check whether the derivative was exact, off by a constant, or missing?
- Did I consciously reject substitution when the structure did not fit?

Being able to say *“this is not a substitution problem”* is a core integration skill.


### Practice: Integration by Parts (Pattern Recognition and Application)

This practice section focuses on **deciding when integration by parts is appropriate** and executing it deliberately.

Remember: integration by parts is not about memorizing a formula—it is about **strategically simplifying a product**.

---

#### Part A: Identifying the Structure (Pattern Recognition) {-}

For each integrand below:

1. Decide whether the integrand is a **product**.
2. Ask the diagnostic question:  
   > *If I differentiate one factor, does it become simpler?*
3. Decide whether integration by parts is appropriate.
4. If yes, identify a **good choice** for \(u\) and \(dv\).

Do **not** compute the integral yet.

---

#### A Helpful Instruction: What “Simpler” Means {-}

A factor becomes *simpler* if differentiation:
- reduces its degree,
- removes structure (e.g., polynomial → constant),
- turns a complicated function into a basic one.

Differentiation that preserves or increases complexity is usually a bad sign.

---

#### Warning: Not Every Product Uses Integration by Parts {-}

Some products are better handled by:
- substitution,
- basic rules,
- or algebraic simplification.

These problems intentionally mix cases where integration by parts **does** and **does not** work.

---

#### A1. {-}

\[
\int x e^x \, dx
\]

#### A2.{-}

\[
\int x \sin x \, dx
\]

#### A3. {-}

\[
\int e^{x^2} \cdot 2x \, dx
\]

#### A4.{-}

\[
\int \ln x \, dx
\]

#### A5. {-}

\[
\int (x + 1)^4 \, dx
\]

#### A6.{-}

\[
\int x^2 e^x \, dx
\]

---

<details>
<summary><strong>Click to reveal analysis</strong></summary>

- **A1**
  - Product: yes  
  - Simplifies when differentiating \(x\)  
  - Method: integration by parts  
  - Good choice: \(u = x\), \(dv = e^x dx\)

- **A2**
  - Product: yes  
  - Simplifies when differentiating \(x\)  
  - Method: integration by parts  
  - Good choice: \(u = x\), \(dv = \sin x dx\)

- **A3**
  - Product: yes  
  - Better structure: nesting  
  - Method: substitution  
  - Reason: derivative matches inner function

- **A4**
  - Product: implicit (\(\ln x \cdot 1\))  
  - Simplifies when differentiating \(\ln x\)  
  - Method: integration by parts  
  - Good choice: \(u = \ln x\), \(dv = dx\)

- **A5**
  - Not a meaningful product  
  - Method: basic power rule  
  - Reason: no simplification via differentiation

- **A6**
  - Product: yes  
  - Simplifies when differentiating \(x^2\)  
  - Method: integration by parts (possibly repeated)

</details>

---

#### Part B: Solve Using Integration by Parts {-}

Now compute each integral **where integration by parts is appropriate**.  
Write your choices of \(u\) and \(dv\) clearly before applying the formula.

---

#### B1.{-}

\[
\int x e^x \, dx
\]

#### B2.{-}

\[
\int x \sin x \, dx
\]

#### B3.{-}

\[
\int \ln x \, dx
\]

#### B4.{-}

\[
\int x^2 e^x \, dx
\]

---

<details>
<summary><strong>Click to reveal solutions</strong></summary>

- **B1**
  - Let \(u = x\), \(dv = e^x dx\)
  \[
  x e^x - e^x + C
  \]

- **B2**
  - Let \(u = x\), \(dv = \sin x dx\)
  \[
  -x \cos x + \sin x + C
  \]

- **B3**
  - Let \(u = \ln x\), \(dv = dx\)
  \[
  x \ln x - x + C
  \]

- **B4**
  - Let \(u = x^2\), \(dv = e^x dx\)

  Apply integration by parts twice:
  \[
  x^2 e^x - 2x e^x + 2e^x + C
  \]

</details>

---

#### Reflection Prompt {-}

After completing this section, ask yourself:

- Did I confirm the integrand was a product *before* choosing integration by parts?
- Did differentiating \(u\) actually simplify the expression?
- If integration by parts failed, did that point me toward a better strategy?

Being able to reject integration by parts confidently is just as important as using it correctly.

### Practice: Partial Fractions (Structured Practice)

This set gives you **two problems of each major type**, with solutions that show the **key algebraic steps**, not just the final answer.

For each problem:
1. Set up the decomposition.
2. Solve for the coefficients.
3. Integrate term by term.

---

#### Type I: Distinct Linear Factors {-}

---

#### P1.1 {-}
 
\[
\int \frac{3x + 5}{(x + 1)(x + 2)}\,dx
\]


#### P1.2{-}

\[
\int \frac{4x - 1}{(x - 1)(x + 3)}\,dx
\]

---

<details>
<summary><strong>Click to reveal solutions</strong></summary>

**P1.1**

Set up:
\[
\frac{3x + 5}{(x + 1)(x + 2)}
=
\frac{A}{x + 1} + \frac{B}{x + 2}.
\]

Clear denominators:
\[
3x + 5 = A(x + 2) + B(x + 1).
\]

Expand:
\[
3x + 5 = (A + B)x + (2A + B).
\]

Match coefficients:
\[
A + B = 3,\quad 2A + B = 5.
\]

Subtract:
\[
A = 2,\quad B = 1.
\]

Integrate:
\[
\int \left(\frac{2}{x + 1} + \frac{1}{x + 2}\right)dx
=
2\ln|x + 1| + \ln|x + 2| + C.
\]



**P1.2**

Set up:
\[
\frac{4x - 1}{(x - 1)(x + 3)}
=
\frac{A}{x - 1} + \frac{B}{x + 3}.
\]

Clear denominators:
\[
4x - 1 = A(x + 3) + B(x - 1).
\]

Expand:
\[
4x - 1 = (A + B)x + (3A - B).
\]

Match coefficients:
\[
A + B = 4,\quad 3A - B = -1.
\]

Add:
\[
4A = 3 \Rightarrow A = \frac{3}{4}.
\]

Then:
\[
B = 4 - \frac{3}{4} = \frac{13}{4}.
\]

Integrate:
\[
\frac{3}{4}\ln|x - 1| + \frac{13}{4}\ln|x + 3| + C.
\]

</details>

---

#### Type II: Repeated Linear Factors {-}

---

#### P2.1 {-}

\[
\int \frac{2x + 3}{(x - 1)^2}\,dx
\]



#### P2.2{-}

\[
\int \frac{x + 1}{(x + 2)^2}\,dx
\]

---

<details>
<summary><strong>Click to reveal solutions</strong></summary>

**P2.1**

Set up:
\[
\frac{2x + 3}{(x - 1)^2}
=
\frac{A}{x - 1} + \frac{B}{(x - 1)^2}.
\]

Clear denominators:
\[
2x + 3 = A(x - 1) + B.
\]

Choose \(x = 1\):
\[
2(1) + 3 = B \Rightarrow B = 5.
\]

Substitute back:
\[
2x + 3 = A(x - 1) + 5 \Rightarrow A = 2.
\]

Integrate:
\[
\int \left(\frac{2}{x - 1} + \frac{5}{(x - 1)^2}\right)dx
=
2\ln|x - 1| - \frac{5}{x - 1} + C.
\]

---

**P2.2**

Set up:
\[
\frac{x + 1}{(x + 2)^2}
=
\frac{A}{x + 2} + \frac{B}{(x + 2)^2}.
\]

Clear denominators:
\[
x + 1 = A(x + 2) + B.
\]

Choose \(x = -2\):
\[
-1 = B.
\]

Substitute back:
\[
x + 1 = A(x + 2) - 1 \Rightarrow A = 1.
\]

Integrate:
\[
\int \left(\frac{1}{x + 2} - \frac{1}{(x + 2)^2}\right)dx
=
\ln|x + 2| + \frac{1}{x + 2} + C.
\]

</details>

---

#### Type III: Irreducible Quadratic Factors {-}

---

#### P3.1 {-}

\[
\int \frac{2x + 1}{x^2 + 1}\,dx
\]

---

#### P3.2{-}

\[
\int \frac{3x}{x^2 + 4}\,dx
\]

---

<details>
<summary><strong>Click to reveal solutions</strong></summary>

**P3.1**

Split:
\[
\int \frac{2x}{x^2 + 1}dx + \int \frac{1}{x^2 + 1}dx.
\]

First term (substitution):
\[
u = x^2 + 1,\quad du = 2x\,dx
\Rightarrow \ln(x^2 + 1).
\]

Second term:
\[
\int \frac{1}{x^2 + 1}dx = \arctan x.
\]

Final:
\[
\ln(x^2 + 1) + \arctan x + C.
\]

---

**P3.2**

Let:
\[
u = x^2 + 4,\quad du = 2x\,dx.
\]

Rewrite:
\[
\int \frac{3x}{x^2 + 4}dx
=
\frac{3}{2}\int \frac{1}{u}du.
\]

Integrate:
\[
\frac{3}{2}\ln(x^2 + 4) + C.
\]

</details>

---

#### Strategy Reminder {-}

- **Linear factors → logarithms**
- **Repeated factors → powers + logarithms**
- **Quadratics → substitution and arctangent**

Partial fractions is successful when the algebra is used to *manufacture familiar antiderivatives*.


### Mixed Practice: Choosing and Executing the Right Method

Real integration problems rarely announce which technique they need.  
This mixed practice set is designed to help you **pause, diagnose, and choose deliberately** before computing.

---

#### Section 1: Which Method? {-}

For each integral below:

1. Decide which method you would use  
   (basic rules, substitution, integration by parts, partial fractions).
2. Give a **one-phrase justification**.
3. Do **not** compute yet.

---

#### M1. {-}
\[
\int x^5 \, dx
\]

#### M2. {-}
\[
\int (3x^2 + 1)^4 \cdot 6x \, dx
\]

#### M3. {-}
\[
\int x e^x \, dx
\]

#### M4. {-}
\[
\int \frac{3x + 2}{(x - 1)(x + 2)} \, dx
\]

#### M5. {-}
\[
\int \cos(4x)\, dx
\]

#### M6. {-}
\[
\int \ln x \, dx
\]

#### M7. {-}
\[
\int \frac{2x}{x^2 + 5} \, dx
\]

#### M8. {-}
\[
\int x^2 e^x \, dx
\]

#### M9. {-}
\[
\int \frac{1}{(x - 3)^2} \, dx
\]

#### M10. {-}
\[
\int (x + 1)^7 \, dx
\]

---

<details>
<summary><strong>Click to reveal method choices</strong></summary>

- **M1**: Basic power rule  
- **M2**: Substitution (nested function + derivative present)  
- **M3**: Integration by parts (product; \(x\) simplifies when differentiated)  
- **M4**: Partial fractions (rational function with factorable denominator)  
- **M5**: Substitution (inner function off by a constant)  
- **M6**: Integration by parts (implicit product \(\ln x \cdot 1\))  
- **M7**: Substitution (numerator is derivative of denominator up to constant)  
- **M8**: Integration by parts (product; requires repetition)  
- **M9**: Basic power rule (rewrite with negative exponent)  
- **M10**: Basic power rule (simple polynomial)

</details>

---

#### Section 2: Solve  {-}

Now compute each integral using the method you identified above.  
Write your setup clearly before integrating.

---

#### M1. {-}
\[
\int x^5 \, dx
\]

#### M2. {-}
\[
\int (3x^2 + 1)^4 \cdot 6x \, dx
\]

#### M3. {-}
\[
\int x e^x \, dx
\]

#### M4. {-}
\[
\int \frac{3x + 2}{(x - 1)(x + 2)} \, dx
\]

#### M5. {-}
\[
\int \cos(4x)\, dx
\]

#### M6. {-}
\[
\int \ln x \, dx
\]

#### M7. {-}
\[
\int \frac{2x}{x^2 + 5} \, dx
\]

#### M8. {-}
\[
\int x^2 e^x \, dx
\]

#### M9. {-}
\[
\int \frac{1}{(x - 3)^2} \, dx
\]

#### M10. {-}
\[
\int (x + 1)^7 \, dx
\]

---

<details>
<summary><strong>Click to reveal solutions</strong></summary>

- **M1**
\[
\frac{x^6}{6} + C
\]

- **M2**
Let \(u = 3x^2 + 1,\; du = 6x\,dx\)
\[
\frac{u^5}{5} + C
=
\frac{(3x^2 + 1)^5}{5} + C
\]

- **M3**
\[
x e^x - e^x + C
\]

- **M4**
Decompose:
\[
\frac{3x + 2}{(x - 1)(x + 2)}
=
\frac{5}{3(x - 1)} + \frac{4}{3(x + 2)}
\]
\[
\frac {5}{3\ln|x - 1|} + \frac{4}{3\ln|x + 2|} + C
\]

- **M5**
\[
\frac{1}{4}\sin(4x) + C
\]

- **M6**
\[
x \ln x - x + C
\]

- **M7**
Let \(u = x^2 + 5,\; du = 2x\,dx\)
\[
\ln(x^2 + 5) + C
\]

- **M8**
Apply integration by parts twice:
\[
x^2 e^x - 2x e^x + 2e^x + C
\]

- **M9**
\[
-\frac{1}{x - 3} + C
\]

- **M10**
\[
\frac{(x + 1)^8}{8} + C
\]

</details>

---

#### Final Reflection  {-}

After finishing, ask yourself:

- Did I pause to identify structure before computing?
- Did my chosen method *simplify* the problem?
- Could I explain my choice in one clear sentence?

Being able to **choose well** is the difference between memorizing techniques and actually doing calculus.


---

### Applications: Choosing and Applying a Technique in an Environmental Context {-}

The problems below come from environmental modeling contexts. For each one, **identify which integration technique applies**, then carry it out. The point isn't just to compute — it's to practice the *recognition step* that decides everything else.

---

#### Applied 1: Soil Moisture Loss After Irrigation {-}

A field plot has just been irrigated. The instantaneous rate at which water is leaving the topsoil (through a combination of evaporation and drainage) is modeled as
\[
\frac{dm}{dt} = -100\,e^{-0.05 t} \quad \text{(g/hr)},
\]
where \(t\) is hours since irrigation ended. Find the **total water lost** between \(t = 0\) and \(t = 24\) hours.

<details>
<summary><strong>Click to reveal solution</strong></summary>

**Technique:** Substitution (with \(u = -0.05 t\), \(du = -0.05\,dt\)). Equivalently, the antiderivative of \(e^{kt}\) is \(\frac{1}{k}e^{kt}\) from the dictionary.

\[
\Delta m = \int_0^{24} -100\,e^{-0.05 t}\,dt
= -100 \cdot \frac{1}{-0.05}\bigl[e^{-0.05 t}\bigr]_0^{24}
= 2000\bigl(e^{-1.2} - 1\bigr).
\]

Numerically, \(e^{-1.2} \approx 0.3012\), so
\[
\Delta m \approx 2000(0.3012 - 1) \approx -1398 \text{ g}.
\]

The negative sign means water *left* the plot. About **1.4 kg of water per unit area** was lost over the first 24 hours.

</details>

---

#### Applied 2: Pollutant Decay in a Small Pond {-}

A pulse of pollutant enters a small pond and degrades over time according to the rate
\[
\frac{dC}{dt} = -\frac{50}{(0.1 t + 1)^2} \quad \text{(mg/L per day)},
\]
where \(t\) is days since the pulse. Find the **net change in concentration** between \(t = 0\) and \(t = 9\) days.

<details>
<summary><strong>Click to reveal solution</strong></summary>

**Technique:** Substitution. Let \(u = 0.1 t + 1\), so \(du = 0.1\,dt\), or \(dt = 10\,du\). When \(t = 0\), \(u = 1\); when \(t = 9\), \(u = 1.9\).

\[
\Delta C = \int_0^9 -\frac{50}{(0.1 t + 1)^2}\,dt
= -50 \cdot 10 \int_1^{1.9} \frac{du}{u^2}
= -500\left[-\frac{1}{u}\right]_1^{1.9}
= -500\left(-\frac{1}{1.9} + 1\right)
\approx -500(0.4737)
\approx -236.8 \text{ mg/L}.
\]

The pond's pollutant concentration *decreased* by about **236.8 mg/L** over the nine-day window.

</details>

---

#### Applied 3: Cumulative Microbial Respiration After a Substrate Pulse {-}

A pulse of glucose is added to a soil microcosm at \(t = 0\). The instantaneous CO₂ respiration rate of the microbial community is modeled as
\[
R(t) = t\,e^{-t/2} \quad \text{(μg C per hour)},
\]
where \(t\) is hours since the pulse. Find the **total carbon respired** between \(t = 0\) and \(t = 6\) hours.

<details>
<summary><strong>Click to reveal solution</strong></summary>

**Technique:** Integration by parts — the integrand is a product where one factor (\(t\)) simplifies under differentiation and the other (\(e^{-t/2}\)) is easy to integrate.

Let \(u = t\), \(dv = e^{-t/2}\,dt\). Then \(du = dt\), \(v = -2 e^{-t/2}\).

\[
\int t\,e^{-t/2}\,dt = -2 t e^{-t/2} - \int -2 e^{-t/2}\,dt
= -2 t e^{-t/2} - 4 e^{-t/2} + C
= -2 e^{-t/2}(t + 2) + C.
\]

Evaluate on \([0, 6]\):
\[
\bigl[-2 e^{-t/2}(t + 2)\bigr]_0^6
= -2 e^{-3}(8) - \bigl(-2 \cdot 1 \cdot 2\bigr)
= -16 e^{-3} + 4.
\]

Numerically, \(e^{-3} \approx 0.0498\), so
\[
\text{Total } \approx -16(0.0498) + 4 \approx 3.20 \text{ μg C}.
\]

About **3.2 μg of carbon** was respired by the community during the six-hour window after the glucose pulse.

</details>

---

#### Applied 4: Heat Lost from a Cooling Storage Tank {-}

A water storage tank cools after being filled with warm water. The instantaneous rate at which heat is lost to the surroundings is
\[
H(t) = (10 - t)\,e^{-t/2} \quad \text{(J/s)},
\]
where \(t\) is seconds since filling. Find the **total heat lost** between \(t = 0\) and \(t = 10\) seconds.

<details>
<summary><strong>Click to reveal solution</strong></summary>

**Technique:** Integration by parts. Let \(u = 10 - t\), \(dv = e^{-t/2}\,dt\). Then \(du = -dt\), \(v = -2 e^{-t/2}\).

\[
\int (10 - t)\,e^{-t/2}\,dt
= -2(10 - t) e^{-t/2} - \int -2 e^{-t/2} \cdot (-1)\,dt
= -2(10 - t) e^{-t/2} - 2 e^{-t/2} + C.
\]

Simplify:
\[
= -2 e^{-t/2}\bigl[(10 - t) + 1\bigr] + C
= -2 e^{-t/2}(11 - t) + C.
\]

Evaluate on \([0, 10]\):
\[
\bigl[-2 e^{-t/2}(11 - t)\bigr]_0^{10}
= -2 e^{-5}(1) - (-2 \cdot 1 \cdot 11)
= -2 e^{-5} + 22.
\]

Numerically, \(e^{-5} \approx 0.00674\), so
\[
\text{Total } \approx -0.013 + 22 \approx 21.99 \text{ J}.
\]

About **22 J** of heat is lost during the first ten seconds.

</details>

---

#### Applied 5: Two-Compartment Nutrient Transfer {-}

After separating variables in a simple two-compartment nutrient-transfer model (soil ↔ root), the cumulative transfer over an interval is given by
\[
T = \int_0^3 \frac{1}{(t+1)(t+3)}\,dt.
\]
Evaluate \(T\) using partial fractions.

<details>
<summary><strong>Click to reveal solution</strong></summary>

**Technique:** Partial fractions — the denominator has two distinct linear factors and the derivative of the denominator isn't sitting in the numerator.

Decompose:
\[
\frac{1}{(t+1)(t+3)} = \frac{A}{t+1} + \frac{B}{t+3}.
\]

Clear the denominator: \(1 = A(t+3) + B(t+1)\). Plug in \(t = -1\): \(1 = 2A \Rightarrow A = \tfrac{1}{2}\). Plug in \(t = -3\): \(1 = -2B \Rightarrow B = -\tfrac{1}{2}\).

So
\[
\int \frac{dt}{(t+1)(t+3)} = \tfrac{1}{2}\ln|t+1| - \tfrac{1}{2}\ln|t+3| + C
= \tfrac{1}{2}\ln\left|\frac{t+1}{t+3}\right| + C.
\]

Evaluate on \([0, 3]\):
\[
T = \tfrac{1}{2}\left[\ln\left|\tfrac{4}{6}\right| - \ln\left|\tfrac{1}{3}\right|\right]
= \tfrac{1}{2}\left[\ln\tfrac{2}{3} - \ln\tfrac{1}{3}\right]
= \tfrac{1}{2}\ln 2
\approx 0.347.
\]

Cumulative nutrient transfer over the three-time-unit window is **\(\tfrac{1}{2}\ln 2 \approx 0.347\)** (in whatever units the rate had).

</details>

---

#### Applied 6: Choose and Apply — Leaf Photosynthesis Pulse {-}

The instantaneous photosynthesis rate of a single leaf during a brief light pulse is modeled as
\[
P(t) = \frac{6 t}{t^2 + 1} \quad \text{(μmol CO}_2\text{/s)},
\]
where \(t\) is seconds since the pulse began. Find the **total CO₂ taken up** between \(t = 0\) and \(t = 4\) seconds.

**Before computing**, identify which technique applies — and why the other techniques are not the right move here.

<details>
<summary><strong>Click to reveal solution</strong></summary>

**Technique selection:**

- *Partial fractions?* The integrand is a rational function with denominator \(t^2 + 1\), but the numerator is \(6t\), and the derivative of the denominator is \(2t\) — sitting right there (up to a constant factor of 3). That's a substitution signal, not a partial-fractions signal.
- *Integration by parts?* No clean product where one factor simplifies on differentiation.
- *Substitution* with \(u = t^2 + 1\), \(du = 2t\,dt\), so \(6t\,dt = 3\,du\):

\[
\int_0^4 \frac{6t}{t^2 + 1}\,dt = 3 \int_1^{17} \frac{du}{u} = 3\bigl[\ln u\bigr]_1^{17} = 3\ln 17.
\]

Numerically, \(\ln 17 \approx 2.833\), so
\[
\text{Total uptake } \approx 3(2.833) \approx 8.50 \text{ μmol CO}_2.
\]

About **8.5 μmol of CO₂** was taken up during the four-second pulse.

The recognition lesson: a rational integrand where \(Q'(x)\) appears (up to a constant) in the numerator is a **substitution** problem, not a partial-fractions problem.

</details>
