# Techniques of Integration (Indefinite Integrals)

## Purpose of the Chapter

In the previous chapter, you learned what a definite integral *means* (accumulation) and why it can be computed using antiderivatives (the Fundamental Theorem of Calculus).  

This chapter answers the next practical question:

> **How do we actually find antiderivatives when the integrand is not “obvious”?**

Rather than treating integration as a list of formulas, this chapter emphasizes **pattern recognition + strategy**: identify the structure of a function, choose an appropriate method, and interpret the result in context.

By the end of this chapter, you should be able to:

- compute **indefinite integrals** using core rules and techniques,
- recognize when to use **substitution**, **integration by parts**, **partial fractions**, or **numerical methods**,
- connect each technique back to derivative rules (chain rule, product rule),
- interpret results using **units, sign, and context**, not just algebra.

---

## Indefinite Integrals and the Constant of Integration

In the previous chapter, you learned how a **definite integral** measures *net accumulated change over a specific interval*. The bounds told you exactly where accumulation started and stopped, and the result was a **single number** with a clear physical meaning.

In this section, we remove those bounds and ask a different—but closely related—question:

> **What functions have a given rate of change?**

The answer leads us to **indefinite integrals**, antiderivatives, and the essential role of the constant of integration.

---

### From Definite to Indefinite Integrals

A **definite integral**
\[
\int_a^b f(x)\,dx
\]
asks:

> *How much total change occurred between \(x=a\) and \(x=b\)?*

An **indefinite integral**
\[
\int f(x)\,dx
\]
asks something more open-ended:

> *What function (or functions) have derivative equal to \(f(x)\)?*

When the bounds disappear, two important things change:

1. **The result is no longer a number.**  
   An indefinite integral produces a *function*.

2. **The accumulation is no longer anchored.**  
   Without a starting point, we cannot say how much has accumulated—only *how accumulation grows*.

This is why we write:
\[
\int f(x)\,dx = F(x) + C,
\]
where \(F'(x) = f(x)\).

The indefinite integral does not describe a single accumulated quantity.  
It describes an entire **family of possible accumulation functions**.

---

### Integrals as Families of Functions

Consider the rate function:
\[
f(x) = 2x.
\]

Any of the following functions has derivative \(2x\):
\[
x^2,\quad x^2 + 3,\quad x^2 - 10,\quad x^2 + C.
\]

```{r antiderivative-family-example, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}

library(ggplot2)
library(patchwork)

# Domain
x <- seq(-3, 3, length.out = 400)

# Rate function
f <- function(x) 2 * x

# Antiderivatives with different constants
F_base  <- x^2
F_up    <- x^2 + 3
F_down  <- x^2 - 10

# Data frames
df_rate <- data.frame(
  x = x,
  y = f(x)
)

df_anti <- data.frame(
  x = x,
  Base = F_base,
  ShiftUp = F_up,
  ShiftDown = F_down
)

# Left panel: rate function
p_rate <- ggplot(df_rate, aes(x = x, y = y)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    title = "Rate Function  f(x) = 2x",
    x = "x",
    y = "Rate"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# Right panel: family of antiderivatives
p_anti <- ggplot(df_anti, aes(x = x)) +
  geom_line(aes(y = Base), linewidth = 1.2, color = "#0072B2") +
  geom_line(aes(y = ShiftUp), linewidth = 1.2, color = "#D55E00") +
  geom_line(aes(y = ShiftDown), linewidth = 1.2, color = "#009E73") +
  labs(
    title = "Antiderivatives  F(x) = x² + C",
    x = "x",
    y = "Accumulated Value"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

# Combine panels
p_rate | p_anti

```

> Remember from Calc I that a derivative function f(x) describes the slopes of a function. The slopes of the functions on the right are all the same for the same x value. So a constant doesn't impact slope. 

All of these functions increase at the same rate for the same values of \(x\).  
The only difference is **where they start**.

This is the key conceptual shift:

> **An indefinite integral describes how accumulation changes, not how much has accumulated so far.**

The missing information is the starting value—and that missing information is encoded in the constant \(C\).

---

### Antiderivatives and the “+ C”

An **antiderivative** of \(f(x)\) is any function whose derivative is \(f(x)\).

Because derivatives eliminate constants:

Differentiate term by term:
\[
\frac{d}{dx}\big(F(x) + C\big)
= \frac{d}{dx}F(x) + \frac{d}{dx}C.
\]

Since a constant does not change with \(x\),
\[
\frac{d}{dx}C = 0.
\]

Therefore,
\[
\frac{d}{dx}\big(F(x) + C\big)
= F'(x) + 0
= F'(x).
\]

every antiderivative comes with an unavoidable ambiguity.

That ambiguity is not a flaw—it reflects reality.

---

#### Why Antiderivatives Are Not Unique

If \(f(x)\) represents a **rate** (such as carbon flux, water inflow, or population growth), then the *accumulated amount* depends on **initial conditions**:

- How much carbon was already present?
- How much water was in the reservoir at the start?
- What was the initial population size?

Different starting values produce different accumulation curves—but all follow the **same rate of change**.

Mathematically, this is why we must include the constant of integration, \(+C\).

---

::: promptbox

**Environmental Example: Two Lakes with the Same Inflow Rate**

Suppose two lakes experience the **same inflow rate** of water, modeled by
\[
f(t) = 2t \quad \text{(m³/day)},
\]
where \(t\) is time in days.

This rate describes how quickly water is entering each lake at any moment.

Integrating the rate gives an accumulation function:
\[
V(t) = t^2 + C,
\]
where \(V(t)\) is the volume of water in the lake and \(C\) represents the **initial volume**.

Now suppose:

- **Lake A** starts with 100 m³ of water  
- **Lake B** starts with 250 m³ of water  

Both lakes gain water according to the *same rate*, so their accumulation functions have the same shape.  
The only difference is their starting point.

- Lake A: \(V_A(t) = t^2 + 100\)
- Lake B: \(V_B(t) = t^2 + 250\)

Graphically, these curves are **vertical shifts** of one another.

```{r two-lakes-same-rate-different-start, echo=FALSE, message=FALSE, warning=FALSE, fig.width=7, fig.height=4}

library(ggplot2)

# Time domain (days)
t <- seq(0, 10, length.out = 400)

# Lake volume functions
VA <- t^2 + 100   # Lake A
VB <- t^2 + 250   # Lake B

# Data frame
df <- data.frame(
  t = t,
  LakeA = VA,
  LakeB = VB
)

# Plot
ggplot(df, aes(x = t)) +
  geom_line(aes(y = LakeA, color = "Lake A"), linewidth = 1.2) +
  geom_line(aes(y = LakeB, color = "Lake B"), linewidth = 1.2) +
  scale_y_continuous(limits = c(0, 400)) +
  labs(
    title = "Two Lakes with the Same Inflow Rate",
    subtitle = "Same rate of change, different initial volumes",
    x = "Time (days)",
    y = "Water Volume (m3)",
    color = NULL
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(face = "bold"),
    plot.subtitle = element_text(color = "gray40")
  )

```
**Key Insight**

> The rate of change determines the *shape* of accumulation.  
> Initial conditions determine the *vertical position*.

This is why antiderivatives are not unique:  
the mathematics reflects the physical reality that systems can evolve at the same rate while starting from different states.

:::


#### Vertical Shifts and Initial Conditions

The constant \(C\) represents a **vertical shift** of the accumulation function.

Conceptually:

- The *shape* of the curve is determined by the rate \(f(x)\).
- The *vertical position* is determined by initial conditions.

This mirrors environmental modeling:

- Same rainfall rate → same growth pattern of water volume  
- Different initial storage → different total volume at all times

Ignoring \(C\) is equivalent to assuming the system always starts at zero—which is rarely justified unless explicitly stated.

---

#### Why Constants Cancel in Definite Integrals

When computing a definite integral using an antiderivative:
\[
\int_a^b f(x)\,dx = [F(x) + C]_a^b,
\]
the constant cancels:
\[
(F(b) + C) - (F(a) + C) = F(b) - F(a).
\]


This is why definite integrals do **not** require a constant of integration.

A definite integral measures **change between two points**, not absolute level.  
Only differences matter.

Indefinite integrals, by contrast, describe **absolute accumulation**, so the constant matters.



---

### Quick Interpretation Habit

To use indefinite integrals correctly—especially in applied contexts—it helps to adopt two quick interpretation checks.

---

#### Units in Indefinite Integrals

Even without bounds, units still carry meaning.

If:

- \(f(x)\) has units of *rate* (e.g., kg/day),
- and \(x\) has units of *time* (days),

then:
\[
\int f(x)\,dx
\]
has units of **accumulated quantity** (kg).

However, without bounds:

- you do **not** yet know *how much* has accumulated,
- only how accumulation depends on \(x\).

This distinction prevents a common mistake: treating an indefinite integral as a physical total without specifying a reference point.

---

#### Determining \(C\) from a Known Starting Value

In real problems, the constant \(C\) is determined using **initial conditions**.

Example:
- Suppose \(f(t)\) is a rate of change.
- You find an antiderivative \(F(t) + C\).
- You are told that the accumulated quantity equals \(A_0\) at time \(t=t_0\).

Then you solve:
\[
F(t_0) + C = A_0
\]
to determine \(C\).

This step turns a *family* of solutions into a **specific model**.

---

### Key Takeaway

> **Indefinite integrals describe how accumulation grows; the constant of integration encodes where accumulation starts.**

Definite integrals eliminate this ambiguity by focusing on change between two points.  
Indefinite integrals retain it—because real systems depend on initial conditions.

Understanding this distinction is essential before learning any integration technique, because every technique ultimately produces an antiderivative—and every antiderivative represents a family of possible accumulated states.


---

## 2) Reverse Differentiation as a Skill

### 2.1 The “Antiderivative Dictionary”
- Power rule in reverse (including the special case \(x^{-1}\))
- Exponentials and logarithms
- Basic trig antiderivatives

### 2.2 Pattern Recognition Over Memorization
- Reading an integrand as a derivative you’ve seen before
- Common structures that should trigger a known rule

### 2.3 Practice: Fast Recognition Drills
- Short, high-repetition integrals
- “Name the rule” before computing

---

## 3) Substitution: When a Function Is Hiding Inside Another

### 3.1 The Big Idea (Chain Rule in Reverse)
- Recognizing \(f(g(x))g'(x)\)
- Why substitution is a “change of variable,” not a trick

### 3.2 The Mechanics of \(u\)-Substitution
- Choosing \(u\)
- Converting \(dx\) to \(du\)
- Integrate in \(u\), then substitute back

### 3.3 Substitution in Definite Integrals (Optional here, or saved for later)
- Changing bounds instead of substituting back
- Why this can reduce algebra mistakes

### 3.4 Common Pitfalls
- Missing the derivative factor
- Incorrect \(du\)
- Substituting back too early

---

## 4) Integration by Parts: When Two Things Are Multiplying

### 4.1 The Big Idea (Product Rule in Reverse)
- Derivation from \((uv)' = u'v + uv'\)
- Why this method “moves complexity” from one factor to another

### 4.2 The Formula and How to Use It
- \(\int u\,dv = uv - \int v\,du\)
- Picking \(u\) and \(dv\) strategically

### 4.3 Strategy for Choosing \(u\)
- Heuristics (what tends to simplify when differentiated)
- Avoiding “no progress” loops

### 4.4 Repeated Integration by Parts (When Needed)
- When one round isn’t enough
- When it terminates vs when it cycles

---

## 5) Trig Integrals and Trig Substitutions (Optional / Course-Dependent)

### 5.1 Trig Identities That Unlock Integrals
- Pythagorean identities
- Power-reduction / rewrite strategies

### 5.2 Common Trig Integral Patterns
- Products of powers of \(\sin\) and \(\cos\)
- Secant/tangent patterns (if included)

### 5.3 Trig Substitution (If included)
- When radicals suggest circles/triangles
- How substitution changes the integrand
- Back-substitution with triangles

---

## 6) Partial Fractions (Rational Functions)

### 6.1 Recognizing “Rational Function” Structure
- When \( \frac{P(x)}{Q(x)} \) is the form
- Proper vs improper rational functions

### 6.2 Algebra Step: Long Division (If Needed)
- Reducing improper to proper + polynomial part

### 6.3 Decomposition Patterns
- Distinct linear factors
- Repeated linear factors
- Irreducible quadratics (if included)

### 6.4 Integrating the Pieces
- Log forms from linear denominators
- Arctan forms from irreducible quadratics (if included)

### 6.5 Common Pitfalls
- Missing terms in the decomposition
- Algebra errors that cascade

---

## 7) Numerical Integration Revisited (A Reality Check)

### 7.1 When You Don’t Have a Formula
- Discrete measurements and tabular data
- Modeling vs measurement

### 7.2 Trapezoid Rule (and Simpson’s Rule if included)
- What the method assumes about the curve
- Accuracy intuition: curvature + resolution

### 7.3 Comparing Numeric and Analytic Results
- Sanity checks
- When disagreement is informative (not just “error”)

---

## 8) Strategy: How to Choose a Method

### 8.1 The Decision Tree
- Basic rules vs substitution vs parts vs partial fractions vs numerical
- “What do you try first?” checklist

### 8.2 Mixed-Technique Problems
- When a problem requires multiple steps
- How to avoid committing too early

### 8.3 Workflow Habits That Prevent Mistakes
- Rewrite/simplify first
- Check by differentiating
- Units and sign checks (especially for applied setups)

---

## Chapter Wrap-Up

### Key Takeaways
- Integration methods are organized around *structure*
- Each technique is the inverse of a derivative rule
- Interpretation (units, sign, context) is part of correctness

### Practice Sets
- Skills drills (fast technique recognition)
- Mixed-method practice
- Applied accumulation problems (analytic + numeric comparison)

