# Workbook: Approximating an Integral: Understanding Area, Accumulation, and Riemann Sums

---

## Activity 1 — What Does “Area Under a Curve” Mean? {-}

### Activity 1A: Area Under a Constant Function {-}

Consider the function  
\[
f(x) = 3 \quad \text{on } [0,5].
\]

**Prompt (Think–Pair–Share):**

- What does the graph of this function look like?
- How would you describe the “area under the curve” in plain language? 
- Can you calculate the area?

<br><br><br><br><br><br><br><br><br><br><br><br>

What if *f* represented the velocity and *x* represented time 

- What would the area under the curve represent?

\[
v(t) = 3 \quad \text{on } [0,5].
\]

<br><br><br><br><br><br><br>



::: promptbox
**Key takeaway:**  
Summary:

<br><br><br><br><br><br>
:::

---

### Activity 1B: Area Under a Piecewise Linear Function {-}

Consider the following piecewise description (a graph should be provided):

- From \(x = 0\) to \(x = 2\): \(f(x) = 2\)
- From \(x = 2\) to \(x = 5\): a straight line from \(f(2) = 2\) to \(f(5) = 6\)

```{r piecewise-linear-plot, echo=FALSE, warning=FALSE, message=FALSE}
# Segment 1: constant from x = 0 to x = 2
x1 <- seq(0, 2, length.out = 100)
y1 <- rep(2, length(x1))

# Segment 2: linear from (2, 2) to (5, 6)
x2 <- seq(2, 5, length.out = 100)
y2 <- 2 + (6 - 2) / (5 - 2) * (x2 - 2)

# Set up the plotting window
plot(
  NA,
  xlim = c(0, 5),
  ylim = c(0, 7),
  xlab = "x",
  ylab = "f(x)",
  main = "Piecewise Linear Function"
)

# Reference lines
abline(h = 0, col = "gray80", lty = 2)
abline(v = 0, col = "gray80", lty = 2)

# Draw the function
lines(x1, y1, lwd = 2)
lines(x2, y2, lwd = 2)

# Mark the join point
points(2, 2, pch = 19)
```

**Guiding Questions:**

- What is the accumulation of this function between [0 5]?

<br><br><br><br><br><br><br><br><br><br><br><br>

::: promptbox
**Key takeaway:**  
Summary:

<br><br><br><br><br><br>
:::


---

### Activity 1C (Student Extrapolation): Area from Discrete Data {-}

You are given the following table of values:

| \(x\)     | 0 | 1 | 2 | 3 | 4 |
|:---------:|:-:|:-:|:-:|:-:|:-:|
| \(f(x)\)  | 2 | 3 | 5 | 4 | 6 |

**Prompt:**

- In this case no function is given to us. Just the data points.
- Estimate the area from \(x = 0\) to \(x = 4\)?

```{r discrete-data-plot, echo=FALSE, warning=FALSE, message=FALSE}
# Given data
x <- c(0, 1, 2, 3, 4)
y <- c(2, 3, 5, 4, 6)

# Create the plot
plot(
  x, y,
  pch = 15,            # square marker
  cex = 1.6,           # larger size
  col = "darkgreen",   # green color
  xlab = "x",
  ylab = "f(x)",
  main = "Discrete Data Points",
  xlim = c(0, 4),
  ylim = c(0, 7)
)

# Reference lines
abline(h = 0, col = "gray80", lty = 2)
abline(v = 0, col = "gray80", lty = 2)

```

<br><br><br><br><br><br><br><br><br><br><br><br>

::: promptbox
**Reflection:**

- What assumptions are you making?
- How might different assumptions change your estimate?
<br><br><br><br><br><br>

:::

---

### Activity 1D: Area Under a Continuous Function {-}

Consider the function  
\[
f(x) = x^2 \quad \text{on } [0,2].
\]

```{r quadratic-shaded-area, echo=FALSE, warning=FALSE, message=FALSE}
# Function and domain
f <- function(x) x^2
x_plot <- seq(-3, 3, length.out = 400)

# Shaded region bounds
a <- 0
b <- 2
x_shade <- seq(a, b, length.out = 200)

# Plot the function over [-3, 3]
plot(
  x_plot, f(x_plot),
  type = "l",
  lwd = 2,
  xlab = "x",
  ylab = "f(x)",
  main = expression(f(x) == x^2),
  xlim = c(-3, 3),
  ylim = c(0, f(3))
)

# Reference lines
abline(h = 0, col = "gray80", lty = 2)
abline(v = 0, col = "gray80", lty = 2)

# Shade the area under the curve on [0, 2]
polygon(
  c(x_shade, rev(x_shade)),
  c(f(x_shade), rep(0, length(x_shade))),
  col = rgb(0, 0, 1, 0.2),
  border = NA
)

# Re-draw the curve on top for clarity
lines(x_plot, f(x_plot), lwd = 2)

# Mark the bounds a and b
abline(v = a, col = "gray50", lty = 3)
abline(v = b, col = "gray50", lty = 3)

```
**Discussion Prompts:**

- Why can’t we use simple geometry here?
- What makes this harder than the previous examples?
- Ideas on how we could 'approximate this area'?
<br><br><br><br><br><br>



---

## Activity 2 — The Big Picture: Riemann Sums {-}

### Reading the Math {-}

**Prompt**

- Can you decipher what this is telling us to do?
<br><br><br><br><br><br>
\[
\sum_{i=1}^{n} f(x_i)\,\Delta x
\]
<br><br><br><br><br><br>

### Activity 2A: Left-Hand Riemann Sum {-}

We return to the function  
\[
f(x) = x^2 \quad \text{on } [0,2].
\]

Now lets pull together the variables we need

**\( \Delta x \)**



<br><br><br><br><br><br>

**\( f(x_i) \)**

<br><br><br><br><br><br><br><br><br><br><br><br>

**\(\sum_{i=1}^{n} f(x_i)\,\Delta x\)**

<br><br><br><br><br><br><br><br><br><br><br><br>

---

### Activity 2B: Right-Hand and Midpoint Sums {-}

**Prompt:**

- If the left-hand sum samples the left endpoint of each subinterval:
  - Where would the right-hand sum sample?
  - Where would the midpoint sum sample?

<br><br><br><br><br>

**Student Tasks:**
Calculate the right-hand and midpoint Riemann sums for 
\[
f(x) = x^2 \quad \text{on } [0,2].
\]
*[Hint: Create a process, build a table of data points, their function values, then calculate the sum]*

<br><br><br><br><br><br><br><br><br><br><br><br>

---

## Activity 3 — Choosing a Method: Conservative vs Optimistic Estimates {-}

### Discussion Activity {-}

Assume the function is **increasing** on the interval.

```{r riemann-under-over-triangles2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}

library(ggplot2)
library(patchwork)

# Increasing rate function
f_fun <- function(t) 0.2 * t + 1

t <- seq(0, 10, length.out = 500)
curve_df <- data.frame(t = t, f = f_fun(t))

# Interval and partition
a <- 2
b <- 8
n <- 3
dt <- (b - a) / n

base <- data.frame(
  t0 = seq(a, b - dt, by = dt)
)
base$t1 <- base$t0 + dt
base$left  <- base$t0
base$right <- base$t1
base$mid   <- (base$t0 + base$t1) / 2

# Build triangle error regions
make_error_triangles <- function(sample_t, type = c("under", "over")) {
  type <- match.arg(type)
  
  do.call(
    rbind,
    lapply(seq_len(nrow(base)), function(i) {
      
      t_seq <- seq(base$t0[i], base$t1[i], length.out = 50)
      curve_vals <- f_fun(t_seq)
      rect_height <- f_fun(sample_t[i])
      
      if (type == "under") {
        keep <- curve_vals > rect_height
        ymin <- rect_height
        ymax <- curve_vals
        fill <- "under"
      } else {
        keep <- curve_vals < rect_height
        ymin <- curve_vals
        ymax <- rect_height
        fill <- "over"
      }
      
      if (!any(keep)) return(NULL)
      
      data.frame(
        t = t_seq[keep],
        ymin = ymin[keep],
        ymax = ymax[keep],
        fill = fill
      )
    })
  )
}

# Plot builder
make_plot <- function(sample_t, title, rect_fill, error_type, subtitle) {
  
  rects <- data.frame(
    t0 = base$t0,
    t1 = base$t1,
    height = f_fun(sample_t),
    sample_t = sample_t
  )
  
  p <- ggplot(curve_df, aes(x = t, y = f)) +
    geom_line(linewidth = 1) +
    
    # Rectangles
    geom_rect(
      data = rects,
      aes(xmin = t0, xmax = t1, ymin = 0, ymax = height),
      inherit.aes = FALSE,
      fill = rect_fill,
      alpha = 0.5,
      color = "black"
    ) +
    
    # Sampling points
    geom_point(
      data = rects,
      aes(x = sample_t, y = height),
      inherit.aes = FALSE,
      size = 3
    ) +
    
    geom_vline(xintercept = c(a, b), linetype = "dashed") +
    
    labs(
      title = title,
      subtitle = subtitle,
      x = "Time",
      y = "Rate"
    ) +
    
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  # Add error shading ONLY if needed
  if (error_type != "none") {
    
    error_df <- make_error_triangles(sample_t, error_type)
    
    p <- p +
      geom_ribbon(
        data = error_df,
        aes(x = t, ymin = ymin, ymax = ymax, fill = fill),
        inherit.aes = FALSE,
        alpha = 0.6
      ) +
      scale_fill_manual(
        values = c(
          "under" = "#F0E442",
          "over"  = "#D55E00"
        )
      )
  }
  
  p
}

# Build panels
p_left <- make_plot(
  base$left,
  "Left Riemann Sum",
  rect_fill = "#E69F00",
  error_type = "under",
  subtitle = "Yellow regions show underestimation"
)

p_mid <- make_plot(
  base$mid,
  "Midpoint Riemann Sum",
  rect_fill = "grey70",
  error_type = "none",
  subtitle = "Over- and underestimates balance"
)

p_right <- make_plot(
  base$right,
  "Right Riemann Sum",
  rect_fill = "#56B4E9",
  error_type = "over",
  subtitle = "Red regions show overestimation"
)

# Combine panels
p_left | p_mid | p_right


```


**Small-Group Discussion Prompts:**

- Which method gives a conservative estimate?
- Which method gives an optimistic estimate?
- When might overestimating be safer?
- When might underestimating be risky?

<br><br><br><br><br><br><br><br><br><br><br><br>

**Assume the function is decreasing**
```{r riemann-under-over-triangles-decreasing2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}

library(ggplot2)
library(patchwork)

# Decreasing rate function
f_fun <- function(t) -0.2 * t + 3

t <- seq(0, 10, length.out = 500)
curve_df <- data.frame(t = t, f = f_fun(t))

# Interval and partition
a <- 2
b <- 8
n <- 3
dt <- (b - a) / n

base <- data.frame(
  t0 = seq(a, b - dt, by = dt)
)
base$t1 <- base$t0 + dt
base$left  <- base$t0
base$right <- base$t1
base$mid   <- (base$t0 + base$t1) / 2

# Build triangle error regions
make_error_triangles <- function(sample_t, type = c("under", "over")) {
  type <- match.arg(type)
  
  do.call(
    rbind,
    lapply(seq_len(nrow(base)), function(i) {
      
      t_seq <- seq(base$t0[i], base$t1[i], length.out = 50)
      curve_vals <- f_fun(t_seq)
      rect_height <- f_fun(sample_t[i])
      
      if (type == "under") {
        keep <- curve_vals > rect_height
        ymin <- rect_height
        ymax <- curve_vals
        fill <- "under"
      } else {
        keep <- curve_vals < rect_height
        ymin <- curve_vals
        ymax <- rect_height
        fill <- "over"
      }
      
      if (!any(keep)) return(NULL)
      
      data.frame(
        t = t_seq[keep],
        ymin = ymin[keep],
        ymax = ymax[keep],
        fill = fill
      )
    })
  )
}

# Plot builder
make_plot <- function(sample_t, title, rect_fill, error_type, subtitle) {
  
  rects <- data.frame(
    t0 = base$t0,
    t1 = base$t1,
    height = f_fun(sample_t),
    sample_t = sample_t
  )
  
  p <- ggplot(curve_df, aes(x = t, y = f)) +
    geom_line(linewidth = 1) +
    
    geom_rect(
      data = rects,
      aes(xmin = t0, xmax = t1, ymin = 0, ymax = height),
      inherit.aes = FALSE,
      fill = rect_fill,
      alpha = 0.5,
      color = "black"
    ) +
    
    geom_point(
      data = rects,
      aes(x = sample_t, y = height),
      inherit.aes = FALSE,
      size = 3
    ) +
    
    geom_vline(xintercept = c(a, b), linetype = "dashed") +
    
    labs(
      title = title,
      subtitle = subtitle,
      x = "Time",
      y = "Rate"
    ) +
    
    theme_minimal() +
    theme(
      legend.position = "none",
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
  
  if (error_type != "none") {
    
    error_df <- make_error_triangles(sample_t, error_type)
    
    p <- p +
      geom_ribbon(
        data = error_df,
        aes(x = t, ymin = ymin, ymax = ymax, fill = fill),
        inherit.aes = FALSE,
        alpha = 0.6
      ) +
      scale_fill_manual(
        values = c(
          "under" = "#56B4E9",  # blue = missing area
          "over"  = "#D55E00"   # red = extra area
        )
      )
  }
  
  p
}

# Build panels (NOTE: error types flipped)
p_left <- make_plot(
  base$left,
  "Left Riemann Sum",
  rect_fill = "#E69F00",
  error_type = "over",
  subtitle = "Overestimates\n(decreasing rate)"
)

p_mid <- make_plot(
  base$mid,
  "Midpoint Riemann Sum",
  rect_fill = "grey70",
  error_type = "none",
  subtitle = "Over- and underestimates balance"
)

p_right <- make_plot(
  base$right,
  "Right Riemann Sum",
  rect_fill = "#009E73",
  error_type = "under",
  subtitle = "Underestimates\n(decreasing rate)"
)

# Combine panels
p_left | p_mid | p_right

```

**Prompt**

- What about now?

<br><br><br><br><br><br><br><br><br><br><br><br>

---

## Activity 4 — Increasing the Number of Intervals {-}

### Activity 4A (Discussion): Accuracy vs Cost {-}

**Prompts:**
- What happens to the the Riemann summ as \(n\) increases?

<br><br><br><br><br><br><br><br><br><br><br><br>

---

### Activity 4B Big-Picture Bridge to Integrals {-}

**Reflective Prompt:**
> In Calculus I, Newton moved from average rates of change to instantaneous rates by letting \(\Delta x \to 0\).

- What do you think happens if we let the number of rectangles grow without bound?

<br><br><br><br><br><br><br><br><br><br><br><br>


```{r refining-approximation-left-sums2, echo=FALSE, message=FALSE, warning=FALSE, fig.width=10, fig.height=4}

library(ggplot2)
library(patchwork)

# Increasing function
f_fun <- function(x) 0.5 * x + 1

# Interval
a <- 0
b <- 6

# True area (exact integral)
true_area <- integrate(f_fun, a, b)$value

# Helper function to build a plot
make_plot <- function(n_rect, title_suffix) {
  
  dx <- (b - a) / n_rect
  x_left <- seq(a, b - dx, by = dx)
  
  rects <- data.frame(
    x0 = x_left,
    x1 = x_left + dx,
    height = f_fun(x_left)
  )
  
  approx_area <- sum(rects$height * dx)
  
  curve_x <- seq(a, b, length.out = 400)
  curve_df <- data.frame(x = curve_x, y = f_fun(curve_x))
  
  ggplot(curve_df, aes(x = x, y = y)) +
    geom_line(linewidth = 1) +
    geom_rect(
      data = rects,
      aes(xmin = x0, xmax = x1, ymin = 0, ymax = height),
      inherit.aes = FALSE,
      fill = "#E69F00",
      alpha = 0.5,
      color = "black"
    ) +
    labs(
      title = paste0("Left Riemann Sum (", n_rect, " rectangles)"),
      subtitle = paste0(
        "Rectangle area = ", round(approx_area, 2),
        "   |   True area = ", round(true_area, 2)
      ),
      x = "x",
      y = "f(x)"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(face = "bold", hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5)
    )
}

# Build panels
p1 <- make_plot(3, "3 rectangles")
p2 <- make_plot(8, "8 rectangles")
p3 <- make_plot(20, "20 rectangles")

# Combine
p1 | p2 | p3

```

- How might this lead us from Riemann sums to *exact* area?

\[
\sum_{i=1}^{n} f(x_i)\,\Delta x \rightarrow \int_{a}^{b} f(x)\,dx
\]

<br><br><br><br><br><br><br><br><br><br><br><br>

---


## Activity 5 — Signed Area: When the Function Goes Negative {-}

In this final activity, we use a **realistic environmental signal** to build intuition for signed area. Many environmental fluxes naturally oscillate between positive and negative values over time.

---

### Activity 5A (Worked Together): Carbon Flux as a Sinusoidal Process {-}

Consider a short-term model of **net carbon exchange** between a forest and the atmosphere over three days.

- Positive values of \(f(t)\): net **carbon sequestration** (uptake)
- Negative values of \(f(t)\): net **carbon release**

Suppose the net carbon flux (in metric tons of CO\(_2\) per day) is modeled by:
\[
f(t) = 2\sin\!\left(2\pi t\right)
\quad \text{on } [0,3],
\]
where \(t\) is time in days.

```{r carbon-sinusoid-1day, echo=FALSE, warning=FALSE, message=FALSE}
# One-day period sinusoid: period = 1 day
f <- function(t) 2 * sin(2 * pi * t)
t <- seq(0, 3, length.out = 800)

plot(
  t, f(t),
  type = "l",
  lwd = 2,
  xlab = "Time (days)",
  ylab = expression("Net carbon flux (metric tons CO"[2]*"/day)"),
  main = "Net Carbon Exchange",
  xlim = c(0, 3),
  ylim = c(-2.5, 2.5)
)

abline(h = 0, col = "gray60", lty = 2)
abline(v = 0:3, col = "gray80", lty = 3)  # day markers at t = 0,1,2,3

```

**Interpretation prompts (class discussion):**

- Why might carbon flux vary smoothly rather than remain constant?
<br><br><br><br><br>

- At what times is the forest acting as a carbon sink?
<br><br><br><br><br>

- At what times is it acting as a carbon source?
<br><br><br><br><br>

- What do the zeros of the function represent physically?
<br><br><br><br><br>

- Lets do a unit check, what are the units of the area under the curve?
<br><br><br><br><br>

- How much **net carbon** has accumulated over the full three days.
<br><br><br><br><br>

- How would you approximate this with a Riemann Sum?
<br><br><br><br><br><br><br><br><br><br>

```{r carbon-sinusoid-lh-riemann-colored, echo=FALSE, warning=FALSE, message=FALSE}
# One-day period sinusoid (period = 1 day)
f <- function(t) 2 * sin(2 * pi * t)

# Smooth curve for the function
t <- seq(0, 3, length.out = 800)

# Left-hand Riemann sum settings
dx <- 0.1
x_left <- seq(0, 3 - dx, by = dx)
h <- f(x_left)

# Plot the function
plot(
  t, f(t),
  type = "l",
  lwd = 2,
  xlab = "Time (days)",
  ylab = expression("Net carbon flux (metric tons CO"[2]*"/day)"),
  main = "Net Carbon Exchange with Left-Hand Riemann Sum (Δx = 0.1)",
  xlim = c(0, 3),
  ylim = c(-2.5, 2.5)
)

# Reference lines
abline(h = 0, col = "gray60", lty = 2)
abline(v = 0:3, col = "gray80", lty = 3)

# Draw LH Riemann rectangles with color + outline
for (i in seq_along(x_left)) {
  x0 <- x_left[i]
  x1 <- x_left[i] + dx
  
  if (h[i] >= 0) {
    rect(
      x0, 0, x1, h[i],
      col = rgb(0, 0.6, 0, 0.4),     # green fill
      border = "darkgreen",
      lwd = 0.8
    )
  } else {
    rect(
      x0, h[i], x1, 0,
      col = rgb(0.8, 0, 0, 0.4),     # red fill
      border = "darkred",
      lwd = 0.8
    )
  }
}

# Re-draw the curve on top
lines(t, f(t), lwd = 2)

```


- Now with a much smaller delta x
```{r carbon-sinusoid-lh-riemann-colored2, echo=FALSE, warning=FALSE, message=FALSE}
# One-day period sinusoid (period = 1 day)
f <- function(t) 2 * sin(2 * pi * t)

# Smooth curve for the function
t <- seq(0, 3, length.out = 800)

# Left-hand Riemann sum settings
dx <- 0.01
x_left <- seq(0, 3 - dx, by = dx)
h <- f(x_left)

# Plot the function
plot(
  t, f(t),
  type = "l",
  lwd = 2,
  xlab = "Time (days)",
  ylab = expression("Net carbon flux (metric tons CO"[2]*"/day)"),
  main = "Net Carbon Exchange with Left-Hand Riemann Sum (Δx = 0.01)",
  xlim = c(0, 3),
  ylim = c(-2.5, 2.5)
)

# Reference lines
abline(h = 0, col = "gray60", lty = 2)
abline(v = 0:3, col = "gray80", lty = 3)

# Draw LH Riemann rectangles with color + outline
for (i in seq_along(x_left)) {
  x0 <- x_left[i]
  x1 <- x_left[i] + dx
  
  if (h[i] >= 0) {
    rect(
      x0, 0, x1, h[i],
      col = rgb(0, 0.6, 0, 0.4),     # green fill
      border = "darkgreen",
      lwd = 0.8
    )
  } else {
    rect(
      x0, h[i], x1, 0,
      col = rgb(0.8, 0, 0, 0.4),     # red fill
      border = "darkred",
      lwd = 0.8
    )
  }
}

# Re-draw the curve on top
lines(t, f(t), lwd = 2)

```

---


::: promptbox
**Key Idea**

Environmental processes often involve **competing fluxes**:
- Uptake vs release
- Storage vs loss

A definite integral does not measure “area” alone—it measures **net accumulation**.

Signed area is the mathematical tool that allows calculus to track *both magnitude and direction* in real environmental systems.

:::



## Closing Reflection {-}

Riemann sums are not the destination.  
They are the bridge between intuition, geometry, and the formal definition of the integral.

## 2 min write {-}

Bullet 3 takeaways from today


