# Differential Equations





#### Graph 1 {-}

For each of these sketches imagine you are floating on a paddleboard. 

- Make an 'x' anywhere on the graph

Now imagine the grey lines are the current below you

- Sketch the path you would take 
  - Go forwards until you leave the bounds of the graph
  - Go backwards until you leave the bounds of the graph
  
You should now have a single trajectory.

Now repeat this process 2 more times, each time putting your 'x' at a different starting point.



```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 1 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  geom_segment(
    data = grid,
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```

::: promptbox

- What happens to your trajectories?

:::

#### Graph 2 {-}

For this graph place 4 *x's* on the left edge of the page.
Sketch the trajectories you create as you move to the right.

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { r * y * (1 - y / K) }
r <- 2      # growth rate
K <- 2.5      # carrying capacity
# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.25)
y_vals <- seq(-3, 3, by = 0.25)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  geom_segment(
    data = grid,
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-1, 4), ylim = c(-1, 4), expand = FALSE) +
  theme_void()


```

::: promptbox

- What happens if you start below the x axis?
- What happens if you start between y=0 and y=2.5?
- What happens if you state above y=2.5?

- How would describe the y=0 line?
- What about the y= 2.5 line?

:::


#### Graph 3 {-}

For this graph place 4 *x's* on the left edge of the page.
Sketch the trajectories you create as you move to the right.

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { r * y * (1 - y / K) - H}
r <- 2      # growth rate
K <- 3     # carrying capacity
H <- 0.6      # carrying capacity

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.3)
y_vals <- seq(-3, 3, by = 0.3)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  geom_segment(
    data = grid,
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-1, 4), ylim = c(-1, 4), expand = FALSE) +
  theme_void()


```

::: promptbox

- There was a slight change between Graph 2 and Graph 3. Can you describe what happened?

:::

#### Graph 4 {-}

For this graph place 4 *x's* on the left edge of the page.
Sketch the trajectories you create as you move to the right.

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { sin(x) - y}
r <- 2      # growth rate
K <- 2.5      # carrying capacity
# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-6, 6, by = 0.4)
y_vals <- seq(-6, 6, by = 0.4)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -5:5
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -5:5, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -5:5, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  geom_segment(
    data = grid,
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-6, 6), ylim = c(-3.5, 3.5), expand = FALSE) +
  theme_void()


```

::: promptbox

- what happens to all the trajectories by the time to get to the right?
- try and plot a line through the field were the grey lines have slope of zero (you may need to interpolate a little)

:::

#### Graph 5 {-}
For this graph place 4 *x's* on the left edge of the page.
Sketch the trajectories you create as you move to the right.

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { x-y^3}
r <- 2      # growth rate
K <- 2.5      # carrying capacity
# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-6, 6, by = 0.4)
y_vals <- seq(-6, 6, by = 0.4)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -5:5
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -5:5, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -5:5, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  geom_segment(
    data = grid,
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-6, 6), ylim = c(-3.5, 3.5), expand = FALSE) +
  theme_void()


```

::: promptbox

- what happens to all the trajectories by the time to get to the right?
- try and plot a line through the field were the grey lines have slope of zero (you may need to interpolate a little)

:::

#### Graph 6 {-}
For this graph place 4 *x's* on the top edge of the page.
Sketch the trajectories you create as you move to the right.


```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 1 - x^2 -y^2}
r <- 2      # growth rate
K <- 2.5      # carrying capacity
# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-6, 6, by = 0.2)
y_vals <- seq(-6, 6, by = 0.2)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -5:5
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -5:5, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -5:5, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  geom_segment(
    data = grid,
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  theme_void()


```

::: promptbox

- what happens to all the trajectories as you move in this field?
- try and plot a line through the field were the grey lines have slope of zero (you may need to interpolate a little)

:::

#### Graph 7 {-}
For this graph place 4 *x's* on the top edge of the page and 4 *x's* on the left edge.
Sketch the trajectories you create as you move to the right.

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ------------------------------------------------------------
# Spiral-in system (phase-plane direction field)
#   x' = -y - a x
#   y' =  x - a y
# This produces trajectories spiraling INTO the origin for a > 0.
# We plot the direction field using dy/dx = (y')/(x').
# ------------------------------------------------------------
a <- 0.4  # damping (increase for faster spiral-in)

dxdt <- function(x, y) { -y - a*x +x^2}
dydt <- function(x, y) {  x - a*y }

# dy/dx for drawing short slope segments in the x-y plane
f <- function(x, y) { dydt(x, y) / dxdt(x, y) }

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.2)
y_vals <- seq(-3, 3, by = 0.2)
grid <- expand.grid(x = x_vals, y = y_vals)

# Avoid division-by-zero / near-vertical issues where dx/dt ~ 0
eps <- 0.08
grid$den <- dxdt(grid$x, grid$y)
grid$slope <- ifelse(abs(grid$den) < eps, NA, f(grid$x, grid$y))

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Direction field (light grey) — omit NAs (near dx/dt = 0)
  geom_segment(
    data = subset(grid, !is.na(slope)),
    aes(x = x - dx/2, y = y - dy/2, xend = x + dx/2, yend = y + dy/2),
    color = "grey70",
    linewidth = 0.5
  ) +

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-3, 3), ylim = c(-3, 3), expand = FALSE) +
  theme_void()
```

::: promptbox

- what happens to all the trajectories by the time to get to the right?
- try and plot a line through the field were the grey lines have slope of zero (you may need to interpolate a little)
- how would you describe (0,0)?

:::


::: promptbox

**SUMMARY**

- What do the grey lines represent?
<br><br>
- What does one of the trajectories represent?
<br><br>
- Why does trajectory vary?
<br><br>
- What do the different *x's* represent?
<br><br>
- What are zero slope grey lines present?
<br><br>
- What does convergence indicate?
<br><br>
- What does divergence indicate?
<br><br>

:::

## Building our own slope field

### Example 2 {-}

The slope field is a way to represent how a point is changing instantaneously.
We can create a 'field' by calculating the value of the derivative on a grid.


Let's consider 
\[ 
f'(x) = 2
\]

If I know the derivative of the function is 2 everywhere I can create a slope field 


```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 0 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```


If we take the integration of f'(x)

\[ 
\int f'(x) dx
\]

\[ 
\int 2 dx = 2x+c
\]

So we have a family of functions that solve the differential equation.

Now this differential equation wasn't that interesting as it was just a constant

### Example 2 {-}

Now let's consider 
\[ 
f'(x) = x 
\]

If I know the derivative of the function is *x* everywhere I can create a slope field 


```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 0 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```

If we take the integration of f'(x)

\[ 
\int f'(x) dx
\]

\[ 
\int x dx = \frac{x^2}{2}+c
\]

So we have a family of functions that solve the differential equation.

You know how to take that antiderivative without too much trouble. Another way to do it is by using **Separation of Variables**.
\[ 
\frac{dy}{dx}=x  
\]
Now separate the variables by grouping all terms with *x's* on one side and *y's* on the other. Keep the *dx* and *dy* in the numerators.

\[ 
dy=x . dx  
\]

Now we are saying a small amount of y *dy* is equal to x times a small about of x *dx*.
So lets accumulate these small amounts by taking the integration of both sides.

\[ 
\int dy= \int x . dx  
\]

\[ 
y= \frac{x^2}{2} +c  
\]

We get the same solution but using a slightly different methodology - this methodology know as the **Separation of Variables** becomes very useful when differential equations are functions of themselves.


Now this differential equation is a little more interesting but is still only a function of x.


### Example 3 {-}

Now let's consider introducing y into our derivative. What does this mean - it means our derivative is a function of both x and y
\[ 
f'(x,y) = y 
\]

If I know the derivative of the function is *y* everywhere I can create a slope field 


```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 0 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```

If we take the integration of f'(x,y)

\[ 
\int f'(x,y) dx
\]

This is a case where separation of variables will be needed.


\[ 
\frac{dy}{dx}=y
\]

### Example 4 {-}

Let's see if we can

- plot the slope field
- sketch a couple of potential solutions
- use separation of variables to solve the differential equations


\[ 
1. f'(x,y) = y/x 
\]

\[ 
2. f'(x,y) = x/y 
\]


\[ 
3. f'(x,y) = -x/y 
\]


```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 0 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 0 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```

```{r,echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)

# ---- Differential equation: define dy/dx = f(x,y) here ----
f <- function(x, y) { 0 }   # change as needed

# ---- Build slope-field grid (for segments) ----
x_vals <- seq(-3, 3, by = 0.5)
y_vals <- seq(-3, 3, by = 0.5)
grid <- expand.grid(x = x_vals, y = y_vals)

grid$slope <- f(grid$x, grid$y)

# Segment length normalization so all dashes are the same length
L <- 0.45
grid$dx <- L / sqrt(1 + grid$slope^2)
grid$dy <- grid$slope * grid$dx

# ---- Axis tick labels placed ON the axes (x-axis at y=0, y-axis at x=0) ----
ticks <- -3:3
tick_df_x <- data.frame(x = ticks, y = 0, lab = as.character(ticks))
tick_df_y <- data.frame(x = 0, y = ticks, lab = as.character(ticks))

# Small offset so labels don't sit exactly on the bold lines
x_label_nudge <- -0.18  # move x-axis numbers slightly below y=0
y_label_nudge <- -0.18  # move y-axis numbers slightly left of x=0

ggplot() +
  # Integer grid
  geom_vline(xintercept = -3:3, color = "grey85", linewidth = 0.4) +
  geom_hline(yintercept = -3:3, color = "grey85", linewidth = 0.4) +

  # Slope field (light grey)
  

  # Bold coordinate axes
  geom_hline(yintercept = 0, color = "black", linewidth = 1.2) +
  geom_vline(xintercept = 0, color = "black", linewidth = 1.2) +

  # Tick labels ON the axes
  geom_text(data = tick_df_x, aes(x = x, y = y + x_label_nudge, label = lab),
            size = 3.5, color = "black") +
  geom_text(data = tick_df_y, aes(x = x + y_label_nudge, y = y, label = lab),
            size = 3.5, color = "black") +

  # Axis labels "x" and "y"
  annotate("text", x = 3.15, y = 0.18, label = "x", size = 4) +
  annotate("text", x = 0.18, y = 3.15, label = "y", size = 4) +

  coord_fixed(xlim = c(-4, 4), ylim = c(-4, 4), expand = FALSE) +
  theme_void()


```



