# Test 3 Review Material {-}

## Test Instructions {-}
Below are the general instructions for all tests

- You will get a chance to **retake** this. Your **highest** score counts toward your final grade.  
- Follow the guidance for each part and **show all work** for full credit.  
- **Non-graphical calculators** are allowed.  
- **One page (two sides)** of handwritten (or font size 8) notes are allowed.
- The equation sheet from the textbook is included as part of the test pack (it doesn't count as note pages)

## Test Format {-}
You should be able to do the following for this test

- Evaluate Signed/NET integrals
- Calculate the average of a function using an integral
- Calculate the total/geometric area between two curves
- Evaluate double integrals
- Evaluate triple integrals
- Be able to explain what the applications above are doing and interpret them
  - Short answer / Multiple Test
  - Interpret / Sketch 


## Part A — Short Answer (Concepts) {-}

There will not be a short answer section - but there might be short answer interpretations at the end of calculations. 

---

### 1. What does a definite integral represent? {-}

<details>
<summary><em>Hint</em></summary>

Think about accumulation. What are we adding up? What does the sign of the function matter?

</details>

<details>
<summary><em>Suggested Answer</em></summary>

A definite integral represents the net accumulation of a quantity over an interval. Geometrically, it is the signed area between the curve and the \(x\)-axis from \(a\) to \(b\).

</details>

---

### 2. Difference between signed integral and total accumulation? {-}

<details>
<summary><em>Hint</em></summary>

What happens when the function is below the \(x\)-axis? Does it subtract or still add?

</details>

<details>
<summary><em>Suggested Answer</em></summary>

A signed integral allows areas below the axis to subtract from areas above. Total accumulation uses absolute value so all contributions are counted positively.

</details>

---

### 3. Why do we use absolute value for geometric area? {-}

<details>
<summary><em>Hint</em></summary>

Area is always positive in geometry.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

Because geometric area measures size, not direction, and must always be positive, we use absolute value to prevent cancellation.

</details>

---

### 4. What must be true about top minus bottom? {-}

<details>
<summary><em>Hint</em></summary>

What ensures the integrand stays positive?

</details>

<details>
<summary><em>Suggested Answer</em></summary>

The function labeled “top” must be greater than or equal to the bottom function across the entire interval.

</details>

---

### 5. What goes wrong if curves cross and you don’t split? {-}

<details>
<summary><em>Hint</em></summary>

Think about sign changes in \(f(x)-g(x)\).

</details>

<details>
<summary><em>Suggested Answer</em></summary>

If the curves cross, the integrand changes sign, leading to cancellation and an incorrect area value.

</details>

---

### 6. Why find intersection points? {-}

<details>
<summary><em>Hint</em></summary>

What happens at those points?

</details>

<details>
<summary><em>Suggested Answer</em></summary>

Intersection points mark where the top and bottom functions switch roles, allowing us to correctly split the integral.

</details>

---

### 7. How do we split if they intersect at \(x=c\)? {-}

<details>
<summary><em>Hint</em></summary>

Write two integrals.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

\[
\text{Area} = \int_a^c (f-g)\,dx + \int_c^b (g-f)\,dx
\]

</details>

---

### 8. How can you quickly determine which function is on top? {-}

<details>
<summary><em>Hint</em></summary>

Test a value in the interval.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

Evaluate both functions at a convenient test point within the interval and compare their values.

</details>

---

### 9. What does average value mean conceptually? {-}

<details>
<summary><em>Hint</em></summary>

Imagine flattening the function into a constant height.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

The average value is the constant height that would produce the same total area over the interval.

</details>

---

### 10. Why divide by \(b-a\)? {-}

<details>
<summary><em>Hint</em></summary>

Think “total divided by length.”

</details>

<details>
<summary><em>Suggested Answer</em></summary>

We divide by the interval length to convert total accumulation into a per-unit average.

</details>

---

### 11. Geometric interpretation of average value? {-}

<details>
<summary><em>Hint</em></summary>

Picture a rectangle.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

It is the height of a rectangle with base \(b-a\) that has the same area as the region under \(f(x)\).

</details>

---

### 12. What does \(dA\) represent? {-}

<details>
<summary><em>Hint</em></summary>

Tiny what?

</details>

<details>
<summary><em>Suggested Answer</em></summary>

\(dA\) represents a tiny piece of area, typically a small rectangle in the plane.

</details>

---

### 13. What does a double integral do that a single integral cannot? {-}

<details>
<summary><em>Hint</em></summary>

Think dimension.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

A double integral accumulates values over a two-dimensional region, not just along a line.

</details>

---

### 14. Why treat outer variable as constant? {-}

<details>
<summary><em>Hint</em></summary>

Integration rules.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

Because when integrating with respect to one variable, all other variables are treated as constants by definition of partial integration.

</details>

---

### 15. What does \(dV\) represent? {-}

<details>
<summary><em>Hint</em></summary>

Tiny 3D piece.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

\(dV\) represents a tiny volume element, such as a small rectangular box.

</details>

---

### 16. What does a triple integral measure with density? {-}

<details>
<summary><em>Hint</em></summary>

Density times volume.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

It measures total quantity (like mass or population) by accumulating density over a three-dimensional region.

</details>

---

### 17. What does changing order of integration mean geometrically? {-}

<details>
<summary><em>Hint</em></summary>

Different slicing direction.

</details>

<details>
<summary><em>Suggested Answer</em></summary>

It represents slicing the same region in a different direction; the total accumulation remains the same if the region and function are unchanged.

</details>


## Part B — Signed Integrals {-}

### 1. Compute  {-}
\[
\int_{-2}^{4} (2x-5)\,dx.
\]

<details>
<summary><em>Hint</em></summary>

Find an antiderivative \(F(x)\) and evaluate \(F(4)-F(-2)\).

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
\int (2x-5)\,dx=x^2-5x
\]

\[
\left[x^2-5x\right]_{-2}^{4}
=(16-20)-\left(4+10\right)
=-4-14=-18
\]

\[
\boxed{-18}
\]

</details>

---

### 2. Compute  {-}
\[
\int_{1}^{e} \frac{1}{x}\,dx.
\]

<details>
<summary><em>Hint</em></summary>

What is the antiderivative of \(1/x\)?

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
\int \frac{1}{x}\,dx=\ln x
\]

\[
\left[\ln x\right]_{1}^{e}=1-0=1
\]

\[
\boxed{1}
\]

</details>

---

### 3. Compute  {-}
\[
\int_{0}^{3} (x^2-4x+1)\,dx.
\]

<details>
<summary><em>Hint</em></summary>

Integrate term by term.

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
\int (x^2-4x+1)\,dx=\frac{x^3}{3}-2x^2+x
\]

\[
\left[\frac{x^3}{3}-2x^2+x\right]_0^3
=9-18+3=-6
\]

\[
\boxed{-6}
\]

</details>

---

### 4. Compute  {-}
\[
\int_{-1}^{2} (3e^x-2)\,dx.
\]

<details>
<summary><em>Hint</em></summary>

Integrate \(e^x\) and constants separately.

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
\int (3e^x-2)\,dx=3e^x-2x
\]

\[
\left[3e^x-2x\right]_{-1}^{2}
=(3e^2-4)-\left(\frac{3}{e}+2\right)
=3e^2-6-\frac{3}{e}
\]

\[
\boxed{3e^2-6-\frac{3}{e}}
\]

</details>

---

## Part C — Average Value {-}

### 5. Find the average value of  {-}
\[
f(x)=x^2+4x
\]  
on \([-1,2]\).

<details>
<summary><em>Hint</em></summary>

Use \(f_{\text{avg}}=\frac{1}{b-a}\int_a^b f(x)\,dx\).

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
f_{\text{avg}}=\frac{1}{3}\int_{-1}^{2}(x^2+4x)\,dx
\]

\[
\int (x^2+4x)\,dx=\frac{x^3}{3}+2x^2
\]

Integral value \(=9\).

\[
f_{\text{avg}}=\frac{9}{3}=3
\]

\[
\boxed{3}
\]

</details>

---

### 6. Find the average value of  {-}
\[
f(x)=e^{2x}
\]  
on \([0,1]\).

<details>
<summary><em>Hint</em></summary>

Remember the interval length is 1.

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
f_{\text{avg}}=\int_0^1 e^{2x}\,dx
\]

\[
=\left[\frac12 e^{2x}\right]_0^1
=\frac12(e^2-1)
\]

\[
\boxed{\frac{e^2-1}{2}}
\]

</details>

---

## Part D — Area Between Curves {-}

### 7. Find the area between  {-}
\[
f(x)=2x+1
\quad\text{and}\quad
g(x)=x^2
\]  
on \([0,2]\).

<details>
<summary><em>Hint</em></summary>

Determine which function is on top.

</details>

<details>
<summary><em>Worked Solution</em></summary>

Since \(2x+1 > x^2\) on \([0,2]\),

\[
\text{Area}=\int_0^2((2x+1)-x^2)\,dx
\]

\[
\int(-x^2+2x+1)\,dx=-\frac{x^3}{3}+x^2+x
\]

\[
\left[-\frac{x^3}{3}+x^2+x\right]_0^2
=-\frac{8}{3}+4+2=\frac{10}{3}
\]

\[
\boxed{\frac{10}{3}}
\]

</details>

---

### 8. Find the total area between {-}
\[
f(x)=x^2-1
\quad\text{and}\quad
g(x)=0
\]
on \([-2,2]\).

<details>
<summary><em>Hint</em></summary>

Find where \(x^2-1=0\). Split the integral and use symmetry.

</details>

<details>
<summary><em>Worked Solution</em></summary>

Zeros at \(x=\pm1\).

\[
\text{Area}=\int_{-2}^{2}|x^2-1|\,dx
=2\left(\int_0^1(1-x^2)\,dx+\int_1^2(x^2-1)\,dx\right)
\]

\[
\int_0^1(1-x^2)\,dx=\frac{2}{3},
\quad
\int_1^2(x^2-1)\,dx=\frac{4}{3}
\]

\[
2\left(\frac{2}{3}+\frac{4}{3}\right)=4
\]

\[
\boxed{4}
\]

</details>

---

### 9. Find the area between {-}
\[
f(x)=e^x
\quad\text{and}\quad
g(x)=x+1
\]
on \([0,1]\).

<details>
<summary><em>Hint</em></summary>

Check where they intersect and determine which is on top.

</details>

<details>
<summary><em>Worked Solution</em></summary>

They meet at \(x=0\). For \(x>0\), \(e^x > x+1\).

\[
\text{Area}=\int_0^1\left(e^x-(x+1)\right)\,dx
\]

\[
=e^x-\frac{x^2}{2}-x\Big|_0^1
=\left(e-\frac12-1\right)-1
=e-\frac{5}{2}
\]

\[
\boxed{e-\frac{5}{2}}
\]

</details>

---

## Part E — Double Integrals {-}

### 10. Evaluate  {-}
\[
\int_{0}^{1}\int_{-1}^{2}(3x-y)\,dy\,dx.
\]

<details>
<summary><em>Hint</em></summary>

Do the inner integral first.

</details>

<details>
<summary><em>Worked Solution</em></summary>

\[
\int_{-1}^{2}(3x-y)\,dy
=\left[3xy-\frac{y^2}{2}\right]_{-1}^{2}
=9x-\frac{3}{2}
\]

\[
\int_0^1\left(9x-\frac{3}{2}\right)\,dx
=3
\]

\[
\boxed{3}
\]

</details>

---

### 11. Evaluate  {-}
\[
\int_{0}^{2}\int_{0}^{2}(x+2y)\,dy\,dx.
\]

<details>
<summary><em>Hint</em></summary>

Integrate with respect to \(y\) first.

</details>

<details>
<summary><em>Worked Solution</em></summary>

Inner:

\[
\int_0^2(x+2y)\,dy
=\left[xy+y^2\right]_0^2
=2x+4
\]

Outer:

\[
\int_0^2(2x+4)\,dx
=\left[x^2+4x\right]_0^2
=12
\]

\[
\boxed{12}
\]

</details>

---

### 12. Evaluate  {-}
\[
\int_{-1}^{1}\int_{0}^{1}(x^2+y)\,dx\,dy.
\]

<details>
<summary><em>Hint</em></summary>

Integrate with respect to \(x\) first.

</details>

<details>
<summary><em>Worked Solution</em></summary>

Inner:

\[
\int_0^1(x^2+y)\,dx
=\left[\frac{x^3}{3}+yx\right]_0^1
=\frac{1}{3}+y
\]

Outer:

\[
\int_{-1}^{1}\left(\frac{1}{3}+y\right)\,dy
=\frac{2}{3}
\]

\[
\boxed{\frac{2}{3}}
\]

</details>

---

## Part F — Triple Integrals {-}

### 13. Evaluate  {-}
\[
\int_{0}^{1}
\int_{0}^{1}
\int_{0}^{2}
(2x+y+z)\,dz\,dy\,dx.
\]

<details>
<summary><em>Hint</em></summary>

Integrate in order: \(z\), then \(y\), then \(x\).

</details>

<details>
<summary><em>Worked Solution</em></summary>

After integrating step-by-step:

\[
\boxed{5}
\]

</details>

---

### 14. Evaluate  {-}
\[
\int_{0}^{1}
\int_{-1}^{1}
\int_{0}^{3}
xy\,dz\,dx\,dy.
\]

<details>
<summary><em>Hint</em></summary>

Look for symmetry in \(x\).

</details>

<details>
<summary><em>Worked Solution</em></summary>

Inner:

\[
\int_0^3 xy\,dz=3xy
\]

Next:

\[
\int_{-1}^{1}3xy\,dx=3y\int_{-1}^{1}x\,dx=0
\]

\[
\boxed{0}
\]

</details>

---

### 15. Evaluate  {-}
\[
\int_{0}^{2}
\int_{0}^{1}
\int_{0}^{1}
(x+y+z)\,dz\,dx\,dy.
\]

<details>
<summary><em>Hint</em></summary>

Work from inside outward.

</details>

<details>
<summary><em>Worked Solution</em></summary>

Inner:

\[
\int_0^1(x+y+z)\,dz=(x+y)+\frac{1}{2}
\]

Next:

\[
\int_0^1\left(x+y+\frac{1}{2}\right)\,dx
=y+1
\]

Outer:

\[
\int_0^2(y+1)\,dy
=4
\]

\[
\boxed{4}
\]

</details>


## Review Sheet: Evaluating Definite Integrals (Net / Signed Accumulation) {-}

---

### What Is a Definite Integral — and Why Do We Care? {-}

A **definite integral**

\[
\int_a^b f(x)\,dx
\]

measures **net accumulation**.

It answers:

- How much total change occurred?
- How much accumulated over time or space?
- What is the *net* effect of gains and losses?

The key idea:

> A definite integral does **not** measure geometric area unless the function is always positive.  
> It measures **signed area**.

Positive values add.  
Negative values subtract.

---

### Big Idea: The Fundamental Theorem of Calculus {-}

If \(F'(x) = f(x)\), then

\[
\int_a^b f(x)\,dx = F(b) - F(a)
\]

So evaluating a definite integral means:

1. Find an antiderivative.
2. Apply bounds (evaluate at top minus bottom).
3. Simplify.

Always subtract:

\[
F(b) - F(a)
\]

---

#### Worked Example 1 {-}

\[
\int_0^4 (x - 2)\,dx
\]

**Step 1: Antiderivative**

\[
\int (x-2)\,dx=\frac{x^2}{2}-2x
\]

**Step 2: Apply the bounds**

\[
\left[\frac{x^2}{2}-2x\right]_0^4
\]

**Step 3: Plug in the top bound \(x=4\)**

\[
\frac{4^2}{2}-2(4)=\frac{16}{2}-8=8-8=0
\]

**Step 4: Plug in the bottom bound \(x=0\)**

\[
\frac{0^2}{2}-2(0)=0
\]

**Step 5: Subtract (top minus bottom)**

\[
0-0=0
\]

\[
\boxed{0}
\]

✔ Net accumulation: **0**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
f1 <- function(x) x - 2
a <- 0
b <- 4

x_all <- seq(a, b, length.out = 600)
y_all <- f1(x_all)

plot(x_all, y_all,
     type = "l", lwd = 2,
     xlab = "x", ylab = "f(x)",
     main = "Worked Example 1: Signed (Net) Accumulation")
abline(h = 0, lty = 2)

# Shade signed area: below axis (negative) and above axis (positive)
x_neg <- seq(a, 2, length.out = 300)
x_pos <- seq(2, b, length.out = 300)

polygon(c(x_neg, rev(x_neg)),
        c(f1(x_neg), rep(0, length(x_neg))),
        density = 20)

polygon(c(x_pos, rev(x_pos)),
        c(f1(x_pos), rep(0, length(x_pos))),
        density = 20)

abline(v = a, lty = 3)
abline(v = b, lty = 3)
```

---

#### Worked Example 2 (Always Positive Function) {-}

\[
\int_1^3 x^2\,dx
\]

**Step 1: Antiderivative**

\[
\int x^2\,dx=\frac{x^3}{3}
\]

**Step 2: Apply the bounds**

\[
\left[\frac{x^3}{3}\right]_1^3
\]

**Step 3: Plug in the top bound \(x=3\)**

\[
\frac{3^3}{3}=\frac{27}{3}=9
\]

**Step 4: Plug in the bottom bound \(x=1\)**

\[
\frac{1^3}{3}=\frac{1}{3}
\]

**Step 5: Subtract (top minus bottom)**

\[
9-\frac{1}{3}=\frac{27}{3}-\frac{1}{3}=\frac{26}{3}
\]

\[
\boxed{\frac{26}{3}}
\]

✔ Net accumulation (and total area): \(\displaystyle \frac{26}{3}\)

```{r, echo=FALSE, message=FALSE, warning=FALSE}
f2 <- function(x) x^2
a <- 1
b <- 3

x_all <- seq(a, b, length.out = 600)
y_all <- f2(x_all)

plot(x_all, y_all,
     type = "l", lwd = 2,
     xlab = "x", ylab = "f(x)",
     main = "Worked Example 2: Area Under a Positive Function")
abline(h = 0, lty = 2)

polygon(c(x_all, rev(x_all)),
        c(y_all, rep(0, length(x_all))),
        density = 20)

abline(v = a, lty = 3)
abline(v = b, lty = 3)
```

---

#### Worked Example 3 (Rate Interpretation) {-}

Suppose \(R(t)=5-t\) is a rate. Find total change from \(t=0\) to \(t=6\):

\[
\int_0^6 (5-t)\,dt
\]

**Step 1: Antiderivative**

\[
\int (5-t)\,dt = 5t-\frac{t^2}{2}
\]

**Step 2: Apply the bounds**

\[
\left[5t-\frac{t^2}{2}\right]_0^6
\]

**Step 3: Plug in the top bound \(t=6\)**

\[
5(6)-\frac{6^2}{2}=30-\frac{36}{2}=30-18=12
\]

**Step 4: Plug in the bottom bound \(t=0\)**

\[
5(0)-\frac{0^2}{2}=0
\]

**Step 5: Subtract (top minus bottom)**

\[
12-0=12
\]

\[
\boxed{12}
\]

✔ Total change: **12**

```{r, echo=FALSE, message=FALSE, warning=FALSE}
R <- function(t) 5 - t
a <- 0
b <- 6

t_all <- seq(a, b, length.out = 600)
y_all <- R(t_all)

plot(t_all, y_all,
     type = "l", lwd = 2,
     xlab = "t", ylab = "R(t)",
     main = "Worked Example 3: Net Accumulation from a Rate")
abline(h = 0, lty = 2)

# Shade signed area: positive from 0 to 5, negative from 5 to 6
t_pos <- seq(a, 5, length.out = 500)
t_neg <- seq(5, b, length.out = 200)

polygon(c(t_pos, rev(t_pos)),
        c(R(t_pos), rep(0, length(t_pos))),
        density = 20)

polygon(c(t_neg, rev(t_neg)),
        c(R(t_neg), rep(0, length(t_neg))),
        density = 20)

abline(v = a, lty = 3)
abline(v = b, lty = 3)
abline(v = 5, lty = 2)  # where the rate crosses 0
```

---

### Practice Problems {-}

#### Part A — Compute the Net Accumulation {-}

1. \[
\int_0^5 (x - 3)\,dx
\]

2. \[
\int_1^4 (2x)\,dx
\]

3. \[
\int_0^2 (4 - x^2)\,dx
\]

4. \[
\int_0^3 (x^2 - 4)\,dx
\]

---

#### Part B — Think Carefully {-}

5. If a function is below the x-axis on an interval, is the definite integral positive or negative? Why?

6. What is the difference between:
   - total geometric area  
   - net (signed) accumulation  

7. If \(\displaystyle \int_2^6 f(x)\,dx=-8\), what does that tell you about the function’s overall behavior on that interval?

---

<details>
<summary><strong>Answer Key</strong></summary>

**Part A Solutions**

**1.** \(\displaystyle -\frac{5}{2}\)

**2.** \(\displaystyle 15\)

**3.** \(\displaystyle \frac{16}{3}\)

**4.** \(\displaystyle -3\)

**Part B Solutions**

**5.** Negative, because signed area below the axis subtracts from net accumulation.

**6.** Total area ignores sign; net accumulation keeps sign and represents overall change.

**7.** Net accumulation is negative; values below the axis dominate overall.

</details>



## Review Sheet: Area Between Two Curves (Total / Geometric Area) {-}

---

### What Does “Area Between Curves” Mean — and Why Do We Care? {-}

When we compute **area between curves**, we are finding **geometric area** — the size of a region in the plane.

Geometric area is always **positive**.

That means:

> We do **not** allow parts of the region to cancel out.

Unlike signed integrals, it does **not** matter whether the curves are above or below the x-axis. The axis is irrelevant. What matters is how the two functions relate to each other — specifically, the **vertical distance between them**. We are measuring the thickness of the region from the top curve down to the bottom curve, wherever that region sits.

This comes up constantly in applications:

- Total difference between two competing growth rates  
- Total “advantage” of one strategy over another  
- Total deviation of a measured signal from a baseline  
- Total habitat area between two boundary curves  

---

### Big Idea {-}

If you integrate **top minus bottom** over an interval, you get:

- **Area**, *only if* the top curve stays on top the whole time.
- **Net signed area**, if the curves switch order and you don’t split.

So the key skill is knowing **when to split** and how to set the **correct bounds**.

---

### Step-by-Step Checklist {-}

1. **Sketch or compare the functions** (even a rough sketch helps).
2. **Find intersection points** (solve \(f(x)=g(x)\)).
3. Use those intersection points to define **interior bounds**.
4. On each sub-interval, determine which curve is **top** and which is **bottom**.
5. Compute area using one of these safe methods:
   - **Method A:** split into pieces where the top curve stays on top  
   - **Method B:** integrate \(|f-g|\) by taking absolute value piece-by-piece  

---

### Technique 1: “Top Function − Bottom Function” (Split at Intersections) {-}

If \(f\) and \(g\) cross at \(x=c\), then you must split:

\[
\text{Area}=\int_a^c \big(\text{top}-\text{bottom}\big)\,dx
+\int_c^b \big(\text{top}-\text{bottom}\big)\,dx
\]

But the “top” might switch after \(c\).

A quick way to determine the top function on an interval:

- Pick a test point \(x^*\) in the interval and compare \(f(x^*)\) and \(g(x^*)\).

---

### Technique 2: Absolute Value Piece-by-Piece {-}

A very safe method is:

\[
\text{Area}=\int_a^b |f(x)-g(x)|\,dx
\]

But **do not** treat \(|f-g|\) as one algebraic expression unless you know where it changes sign.

In practice, this becomes:

\[
\text{Area}
=
\int_a^c |f-g|\,dx+\int_c^b |f-g|\,dx
\]

On each piece, \(|f-g|\) becomes either \((f-g)\) or \((g-f)\).

---

### Worked Example 1: Curves Do Not Cross Inside the Interval {-}

Find the total area between:

\[
f(x)=2x+1
\qquad
g(x)=x^2
\]

on \([0,2]\).

#### Step 1: Find intersection points {-}

Solve:

\[
2x+1=x^2
\Rightarrow x^2-2x-1=0
\]

\[
x=1\pm \sqrt{2}
\]

Neither solution lies in \([0,2]\), so there are **no interior crossings**.

#### Step 2: Determine top vs bottom {-}

Pick \(x=1\):

\[
f(1)=3,\quad g(1)=1
\Rightarrow f \text{ is on top}
\]

#### Step 3: Set up the area integral (top − bottom) {-}

\[
\text{Area}=\int_0^2 \big((2x+1)-x^2\big)\,dx
\]

#### Step 4: Integrate {-}

\[
\int (-x^2+2x+1)\,dx
=
-\frac{x^3}{3}+x^2+x
\]

#### Step 5: Apply bounds {-}

\[
\left[-\frac{x^3}{3}+x^2+x\right]_0^2
\]

Top bound \(x=2\):

\[
-\frac{8}{3}+4+2=\frac{10}{3}
\]

Bottom bound \(x=0\):

\[
0
\]

Subtract:

\[
\frac{10}{3}-0=\frac{10}{3}
\]

\[
\boxed{\frac{10}{3}}
\]

```{r, echo=FALSE, message=FALSE, warning=FALSE}
f <- function(x) 2*x + 1
g <- function(x) x^2
a <- 0
b <- 2

x <- seq(a, b, length.out = 600)

plot(x, f(x), type="l", lwd=2,
     xlab="x", ylab="y",
     main="Worked Example 1: Area Between Curves",
      xlim = c(-0.2, 2.2),
     ylim = c(-0.5, 5.5))

lines(x, g(x), lwd=2, lty=2)

polygon(c(x, rev(x)),
        c(f(x), rev(g(x))),
        density=20)

abline(v=a, lty=3)
abline(v=b, lty=3)
legend("topleft",
       legend=c("f(x)=2x+1","g(x)=x^2"),
       lty=c(1,2), bty="n")
```

---

### Worked Example 2: Curves Cross Inside the Interval (Must Split) {-}

Find the total area between:

\[
f(x)=x
\qquad
g(x)=x^3
\]

on \([-1,1]\).

#### Step 1: Find intersection points (interior bounds) {-}

Solve:

\[
x=x^3
\Rightarrow x(x^2-1)=0
\Rightarrow x=-1,0,1
\]

So split at:

\[
x=-1,\;0,\;1
\]

#### Step 2: Determine top function on each piece {-}

- On \([-1,0]\): test \(x=-0.5\)

\[
f(-0.5)=-0.5,\quad g(-0.5)=-0.125
\Rightarrow g \text{ is on top}
\]

- On \([0,1]\): test \(x=0.5\)

\[
f(0.5)=0.5,\quad g(0.5)=0.125
\Rightarrow f \text{ is on top}
\]

---

#### Method A: Split and Use Top − Bottom {-}

\[
\text{Area}
=
\int_{-1}^{0} (x^3-x)\,dx
+
\int_{0}^{1} (x-x^3)\,dx
\]

Antiderivative:

\[
\int (x-x^3)\,dx=\frac{x^2}{2}-\frac{x^4}{4}
\]

First piece:

\[
\left[\frac{x^4}{4}-\frac{x^2}{2}\right]_{-1}^{0}
\]

Second piece:

\[
\left[\frac{x^2}{2}-\frac{x^4}{4}\right]_{0}^{1}
\]

Each evaluates to \(\frac14\).

Total area:

\[
\frac14+\frac14=\frac12
\]

\[
\boxed{\frac12}
\]

---

#### Method B: Absolute Value Piece-by-Piece {-}

\[
\text{Area}=\int_{-1}^{1} |x-x^3|\,dx
\]

Split where sign changes:

\[
=
\int_{-1}^{0} |x-x^3|\,dx
+
\int_{0}^{1} |x-x^3|\,dx
\]

On \([-1,0]\), \(x-x^3\le0\), so:

\[
|x-x^3|=x^3-x
\]

On \([0,1]\), \(x-x^3\ge0\), so:

\[
|x-x^3|=x-x^3
\]

This produces the exact same two integrals as Method A.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
f <- function(x) x
g <- function(x) x^3
a <- -1
b <- 1
c <- 0

x_left <- seq(a, c, length.out = 400)
x_right <- seq(c, b, length.out = 400)

plot(seq(a, b, length.out=800), f(seq(a, b, length.out=800)),
     type="l", lwd=2,
     xlab="x", ylab="y",
     main="Worked Example 2: Split Where Curves Cross")
lines(seq(a, b, length.out=800), g(seq(a, b, length.out=800)),
      lwd=2, lty=2)

polygon(c(x_left, rev(x_left)),
        c(g(x_left), rev(f(x_left))),
        density=20)

polygon(c(x_right, rev(x_right)),
        c(f(x_right), rev(g(x_right))),
        density=20)

abline(v=c, lty=3)
abline(v=a, lty=3)
abline(v=b, lty=3)

legend("topleft",
       legend=c("f(x)=x","g(x)=x^3"),
       lty=c(1,2), bty="n")
```

---

### Common Mistakes to Avoid {-}

- Forgetting to find intersection points before integrating  
- Assuming the same function stays on top across the whole interval  
- Forgetting to split when the integrand changes sign  
- Using absolute value without identifying where the sign changes  

---

### Practice Problems {-}

1. Find the total area between:
\[
f(x)=x^2,\quad g(x)=2x
\quad \text{on } [0,2]
\]

2. Find the total area between:
\[
f(x)=x^3,\quad g(x)=x
\quad \text{on } [-1,1]
\]

---

<details>
<summary><strong>Answer Key</strong></summary>

**1.**
\[
\text{Area}=\int_0^2(2x-x^2)\,dx=\frac{4}{3}
\]

**2.**
\[
\text{Area}=\frac12
\]

</details>

## Review Sheet: Double and Triple Integrals {-}

---

### What Do Double and Triple Integrals Represent — and Why Do We Care? {-}

A single integral accumulates along a **line**.

A double integral accumulates over a **region in the plane**.

A triple integral accumulates throughout a **volume in space**.

The idea is the same as with single integrals:

> We are adding up tiny pieces.

But now the tiny pieces are not just little line segments — they are:

- Tiny rectangles (area elements)
- Tiny boxes (volume elements)

---

### Double Integrals — Big Idea {-}

A double integral


\[
\iint_R f(x,y)\, dA
\]


means:

- Break region \(R\) into tiny rectangles \(dA\).
- Evaluate the function at each small rectangle.
- Add everything together.

If \(f(x,y)=1\), then the double integral simply gives **area of the region**.

If \(f(x,y)\) is a density, it gives **total mass**.

If \(f(x,y)\) is height, it gives **volume under a surface**.

---

### How to Evaluate a Double Integral {-}

Most problems are written as **iterated integrals**:

\[
\int_a^b \int_{c(x)}^{d(x)} f(x,y)\,dy\,dx
\]

This means:

1. Integrate the inner integral first (with respect to \(y\)).
2. Treat the outer variable \(x\) as constant.
3. Then integrate the result with respect to \(x\).

Important:

- The inner bounds belong to the inner variable.
- The outer bounds belong to the outer variable.

---

### Worked Example 1 (Rectangular Region) {-}

Evaluate:

\[
\int_0^2 \int_0^1 (3x + 2y)\,dy\,dx
\]

#### Step 1: Integrate with respect to \(y\) {-}

Treat \(x\) as constant.

\[
\int (3x + 2y)\,dy
=
3xy + y^2
\]

Apply bounds:

\[
\left[3xy + y^2\right]_0^1
\]

Top bound \(y=1\):

\[
3x(1) + 1^2 = 3x + 1
\]

Bottom bound \(y=0\):

\[
0
\]

So the inner integral becomes:

\[
3x + 1
\]

---

#### Step 2: Integrate with respect to \(x\) {-}

\[
\int_0^2 (3x + 1)\,dx
=
\left[\frac{3}{2}x^2 + x\right]_0^2
\]

Top bound \(x=2\):

\[
\frac{3}{2}(4) + 2 = 6 + 2 = 8
\]

Bottom bound \(x=0\):

\[
0
\]

Final answer:

\[
\boxed{8}
\]

---

#### Common Mistake (Double Integrals) {-}

Students often forget:

- When integrating with respect to \(y\), \(x\) is treated as constant.
- Bounds must match the variable of integration.

---

### Triple Integrals — Big Idea {-}

A triple integral


\[
\iiint_E f(x,y,z)\, dV
\]


means:

- Break a 3D region into tiny boxes \(dV\).
- Evaluate the function at each box.
- Add them all together.

If \(f(x,y,z)=1\), the result is **volume**.

If \(f\) is density, the result is **total mass**.

---

### How to Evaluate a Triple Integral {-}

Typical form:

\[
\int_a^b
\int_{c(x)}^{d(x)}
\int_{e(x,y)}^{f(x,y)}
f(x,y,z)\,dz\,dy\,dx
\]

You integrate:

1. Inside first.
2. Work outward step-by-step.
3. Treat outer variables as constants each time.

---

### Worked Example 2 (Rectangular Box) {-}

Evaluate:

\[
\int_0^1
\int_0^1
\int_0^2
(2x + y + z)
\,dz\,dy\,dx
\]

#### Step 1: Integrate with respect to \(z\) {-}

Treat \(x\) and \(y\) as constants.

\[
\int (2x + y + z)\,dz
=
(2x + y)z + \frac{z^2}{2}
\]

Apply bounds:

\[
\left[(2x + y)z + \frac{z^2}{2}\right]_0^2
\]

Top bound \(z=2\):

\[
2(2x + y) + 2
=
4x + 2y + 2
\]

Bottom bound \(z=0\):

\[
0
\]

So the expression becomes:

\[
4x + 2y + 2
\]

---

#### Step 2: Integrate with respect to \(y\) {-}

\[
\int_0^1 (4x + 2y + 2)\,dy
=
4xy + y^2 + 2y
\]

Apply bounds:

\[
\left[4xy + y^2 + 2y\right]_0^1
=
4x + 1 + 2
=
4x + 3
\]

---

#### Step 3: Integrate with respect to \(x\) {-}

\[
\int_0^1 (4x + 3)\,dx
=
\left[2x^2 + 3x\right]_0^1
=
2 + 3
=
\boxed{5}
\]

---

### Conceptual Reminders {-}

- A double integral adds over **area**.
- A triple integral adds over **volume**.
- Order of integration can change, but the total remains the same if the region is unchanged.
- Always match bounds to the variable being integrated.
- Always treat outer variables as constants when integrating inside.

---

### Practice Problems {-}

1. Evaluate:
\[
\int_0^1 \int_0^2 (x + y)\,dy\,dx
\]

2. Evaluate:
\[
\int_0^1 \int_0^1 \int_0^1 xyz\,dz\,dy\,dx
\]

3. Evaluate:
\[
\int_{0}^{1}\int_{0}^{1-x}\int_{0}^{1-x-y} (x+2y+3z)\,dz\,dy\,dx
\]

4. What does a double integral compute if \(f(x,y)=1\)?

5. What does a triple integral compute if \(f(x,y,z)=\rho(x,y,z)\)?

---

<details>
<summary><strong>Answer Key</strong></summary>

**1.**  
Inner: \(xy + \frac{y^2}{2}\big|_0^2 = 2x + 2\)  
Outer: \(\int_0^1 (2x+2)\,dx = 3\)

**2.**  
Separate variables:
\[
\frac18
\]

**3.**
\[
\int_{0}^{1}\int_{0}^{1-x}\int_{0}^{1-x-y} (x+2y+3z)\,dz\,dy\,dx
= \frac{1}{4}
\]

**4.** Area of the region.

**5.** Total mass (density integrated over volume).

</details>
